---
title: "GrowthModelFit_Feb16"
author: "Stephen R. Proulx"
date: "2026-03-19"
output:
  pdf_document:
    keep_tex: true
    latex_engine: xelatex
---

```{r knitr_options_growth_fit, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(cmdstanr)
library(posterior)
library(loo)
library(bayesplot)

options(mc.cores = parallel::detectCores())

```

```{r constants_growth_fit}
# --- File paths ---
growth_data_path <- "../cue_data.csv" 

# --- CmdStan sampling defaults ---
growth_chains <- 6
growth_parallel_chains <- 6
growth_iter_warmup <- 2000
growth_iter_sampling <- 4000
growth_adapt_delta <- 0.95
growth_max_treedepth <- 12
```

```{r growth_step1_load_raw}
d_growth_raw <- read_csv(growth_data_path, show_col_types = FALSE)

d_growth_raw %>% summarise(N_raw = n())
```

```{r growth_step2_process_encode}
d_growth <- d_growth_raw %>%
  filter(!is.na(growth_rate)) %>% #remove NA data
  mutate(
    population = recode(
      population,
      "UNREL22"  = "UNREL_22",
      "UNREL120" = "UNREL_120" #fix coding errors in data file: Two pops were inconsistently labelled.
    ),
    Ci = if_else(cue == "light", 1L, -1L),
    is_evo = if_else(EE == "ANC", 0L, 1L),
    evo_type = case_when(
      EE == "ANC"   ~ 0L,
      EE == "REL"   ~ 1L,
      EE == "UNREL" ~ 2L,
      TRUE          ~ NA_integer_
    ),
    g_chr = as.character(grandmother),
    m_chr = as.character(mother),
    mat_code = paste0(g_chr, m_chr),
    mat_id = case_when(
      mat_code == "AN" ~ 1L,
      mat_code == "NN" ~ 2L,
      mat_code == "AA" ~ 3L,
      mat_code == "NA" ~ 4L,
      TRUE ~ NA_integer_
    ),
    cue_index = if_else(Ci == -1L, 1L, 2L),
    state_id = if_else(!is.na(mat_id), 2L * (mat_id - 1L) + cue_index, NA_integer_),
    log_growth = log(growth_rate),
    block_id = as.integer(factor(block))
  ) %>%
  mutate(
    EE_seq_chr = as.character(EE_seq),
    EE_seq_id = if_else(is.na(EE_seq_chr), 0L, as.integer(factor(EE_seq_chr)))
  ) %>%
  filter(!is.na(log_growth))

d_growth %>%
  summarise(
    N = n(),
    n_anc = sum(evo_type == 0L),
    n_rel = sum(evo_type == 1L),
    n_unrel = sum(evo_type == 2L),
    B = n_distinct(block_id),
    K_seq = max(EE_seq_id),
    Ci_vals = paste(sort(unique(Ci)), collapse = ", "),
    mat_vals = paste(sort(unique(mat_id)), collapse = ", "),
    state_range = paste(range(state_id), collapse = " .. ")
  )
```

```{r growth_03_population_ids}
rel_levels <- d_growth %>%
  filter(evo_type == 1L) %>%
  distinct(population) %>%
  arrange(population) %>%
  pull(population)

unrel_levels <- d_growth %>%
  filter(evo_type == 2L) %>%
  distinct(population) %>%
  arrange(population) %>%
  pull(population)

d_growth <- d_growth %>%
  mutate(
    pop_rel_id = if_else(evo_type == 1L, match(population, rel_levels), 0L),
    pop_unrel_id = if_else(evo_type == 2L, match(population, unrel_levels), 0L)
  )

d_growth %>%
  summarise(
    J_rel = max(pop_rel_id),
    J_unrel = max(pop_unrel_id),
    pop_ok =
      !any(evo_type != 1L & pop_rel_id != 0L) &
      !any(evo_type != 2L & pop_unrel_id != 0L)
  )
```

```{r growth_04_sanity_checks}
stopifnot(all(d_growth$Ci %in% c(-1L, 1L)))
stopifnot(all(d_growth$mat_id %in% 1:4))
stopifnot(all(d_growth$state_id %in% 1:8))
stopifnot(all(d_growth$is_evo %in% c(0L, 1L)))
stopifnot(all(d_growth$evo_type %in% c(0L, 1L, 2L)))
stopifnot(all(d_growth$block_id >= 1L))
stopifnot(all(d_growth$EE_seq_id >= 0L))
stopifnot(all(d_growth$pop_rel_id >= 0L))
stopifnot(all(d_growth$pop_unrel_id >= 0L))
```

```{r growth_05_build_stan_data}
stan_data_growth <- list(
  N = nrow(d_growth),
  y = as.vector(d_growth$log_growth),
  is_evo   = as.integer(d_growth$is_evo),
  evo_type = as.integer(d_growth$evo_type),
  Ci       = as.integer(d_growth$Ci),
  mat_id   = as.integer(d_growth$mat_id),
  state_id = as.integer(d_growth$state_id),
  use_state_anc = 1L,
  use_state     = c(1L, 1L),
  B        = max(d_growth$block_id),
  block_id = as.integer(d_growth$block_id),
  K_seq  = max(d_growth$EE_seq_id),
  seq_id = as.integer(d_growth$EE_seq_id),
  J_rel      = max(d_growth$pop_rel_id),
  pop_rel_id = as.integer(d_growth$pop_rel_id),
  J_unrel      = max(d_growth$pop_unrel_id),
  pop_unrel_id = as.integer(d_growth$pop_unrel_id)
)

str(stan_data_growth, max.level = 1)
```


Here is stan code for just the full interaction model. This is new code and has not yet replaced the old code with all the flags and switches. We will set it up to be able to run and compare the total output. 
```{r growth_065_fit_full_interaction}
stan_code_growth_interactionsonly <- "


functions {
  // Center a vector so its elements sum to zero. Used to enforce sum-to-zero effects expressed
  // as deviations from a baseline
  vector center_sum_to_zero(vector x) {
    return x - mean(x);
  }
}

data {
  // Number of observations
  int<lower=1> N;

  // log fitness, already transformed in data 
  vector[N] y;

  // Indicator for whether an observation comes from an evolved population
  // 0 = ancestral, 1 = evolved
  array[N] int<lower=0, upper=1> is_evo;

  // Type of evolved treatment
  // 0 = ancestral 
  // 1 = reliable evolved
  // 2 = unreliable evolved
  array[N] int<lower=0, upper=2> evo_type;

  // State index for each observation
  // Full interaction model uses 8 states that combine blue light cue and maternal/grandmaternal oxygen environment.
  array[N] int<lower=1, upper=8> state_id;

  // Number of assay blocks
  int<lower=1> B;

  // Block identity for each observation
  array[N] int<lower=1, upper=B> block_id;

  // Number of environmental sequences used for evolved treatments
  int<lower=0> K_seq;

  // Sequence identity for each observation
  // We label the Ancestral rows can with seq_id = 0 because they are not used.
  array[N] int<lower=0, upper=K_seq> seq_id;

  // Number of reliable replicate populations
  int<lower=0> J_rel;

  // Reliable population number, within reliable numbered form 1 to j
  // all non-reliable rows are labeled with pop_rel_id = 0 because it won't be used
  array[N] int<lower=0, upper=J_rel> pop_rel_id;

  // Number of unreliable replicate populations
  int<lower=0> J_unrel;

  // Unreliable population number, within unreliable numbered form 1 to j
  // all non-unreliable rows are labeled with pop_unrel_id = 0 because it won't be used
  array[N] int<lower=0, upper=J_unrel> pop_unrel_id;
}

parameters {
  // Overall ancestral intercept
  // This is the average ancestral log fitness before adding state effects
  real alpha_anc;

  // Raw ancestral state effects for the 8 states
  // These will later made to sum to zero and scaled by a sigma term-sigma_state_anc
  vector[8] state_raw_anc;

  // Standard deviation for ancestral state effects
  real<lower=0> sigma_state_anc;

  // Evolved-treatment shifts in intercept relative to the ancestor, relaible 1 unreliable 2
  vector[2] delta_alpha;

  // treatment-specific deviations in state effects
  // These will later made to sum to zero and scaled by a sigma term, relaible 1 unreliable 2
  matrix[2, 8] delta_state_raw;

  // Standard deviations for ancestral state effects, relaible 1 unreliable 2
  vector<lower=0>[2] sigma_delta_state;

  // Raw block effects, will be scaled and sum-to-zero. These apply to all observations in this block, regardless of experimental evolution treatment. 
  vector[B] z_block;

  // Standard deviation for block effects
  real<lower=0> sigma_block;

  // Evolutionary sequence (order of A/N) effects. Only applies to evolved populations. 
  // additive model, applies to both Rel and Unrel with same parameter
  // Will be sum-to-zero and scalled
  vector[K_seq] z_seq;

  // Standard deviation for sequence effects
  real<lower=0> sigma_seq;

  // reliable population effects
  // Will be sum-to-zero and scalled
  vector[J_rel] z_pop_rel;

  // unreliable population effects
  // Will be sum-to-zero and scalled
  vector[J_unrel] z_pop_unrel;

  // standard deviation for replicate population effect
  // This is shared between reliable and unreliable populations, but
  // sum-to-zero acts within each evolutionary treatment. 
  real<lower=0> sigma_pop;

  // standard deviation for the likelihood of observing each data point 
  real<lower=0> sigma_y;
}

transformed parameters {
  // ancestral state effects
  // Apply sigma and sum to zero. 
  vector[8] state_anc = center_sum_to_zero(sigma_state_anc * state_raw_anc);

  // Evolutionary treatment state effects- Reliable, Unreliable
  // Multiply by the appropriate sigma_delta and then enforce sum-to-zero
  matrix[2, 8] delta_state;
  for (t in 1:2) {
    delta_state[t] =
      to_row_vector(center_sum_to_zero(sigma_delta_state[t] * to_vector(delta_state_raw[t])));
  }

  // block effects, apply sigma and sum-to-zero
  vector[B] block_eff = center_sum_to_zero(sigma_block * z_block);

  // Sequence effects, multiply by sigma and sum-to-zero
 vector[K_seq] seq_eff = center_sum_to_zero(sigma_seq * z_seq);

  // Reliable population effects, multiply by sigma and sum-to-zero
  vector[J_rel] pop_rel_eff = center_sum_to_zero(sigma_pop * z_pop_rel);

  // Unreliable population effects, multiply by sigma and sum-to-zero
  vector[J_unrel] pop_unrel_eff = center_sum_to_zero(sigma_pop * z_pop_unrel);

}

model {
  // Prior for ancestral intercept
  alpha_anc ~ normal(0, 1.5);

  // Priors for ancestral state effects
  // Raw effects have standard normal prior
  // Their total scale is controlled by sigma_state_anc
  state_raw_anc ~ normal(0, 1);
  sigma_state_anc ~ exponential(1);

  // Priors for treatment-specific intercept shifts, quite broad but still regularizing
  delta_alpha ~ normal(0, 1);

  // Priors for treatment-specific state deviations
  // Raw deviations are standard normal
  // Their magnitudes are controlled by sigma_delta_state, this is common between
  // Rel and Unrel, but unique from Ancestral.
  to_vector(delta_state_raw) ~ normal(0, 1);
  sigma_delta_state ~ exponential(1);

  // Priors for block  effects
  z_block ~ normal(0, 1);
  sigma_block ~ exponential(1);

  // Priors for sequence  effects
    z_seq ~ normal(0, 1);
    sigma_seq ~ exponential(1);

  // Priors for replicate population effects
    z_pop_rel ~ normal(0, 1);
    z_pop_unrel ~ normal(0, 1);
  
  // Shared prior scale for replicate population effects
  sigma_pop ~ exponential(1);

  // Prior for normal likelihood standard deviation
  sigma_y ~ exponential(1);

  // Likelihood starts here
  for (n in 1:N) {
    // divide into fixed and non-fixed effects
    real mu_fixed;

    // other contributions, labelled as epsilons in print version
    real re_block;
    real re_seq;
    real re_pop;

    // For ancestral observations:
    // mu = ancestral intercept + ancestral state effect
    if (is_evo[n] == 0) {
      mu_fixed = alpha_anc + state_anc[state_id[n]];
    } else {
      // For evolved observations we have to also include shift in intercept and shift in 
      // state-specific response.
      int t = evo_type[n];
      mu_fixed =
        (alpha_anc + delta_alpha[t]) +
        (state_anc[state_id[n]] + delta_state[t, state_id[n]]);
    }

    // Block effect, each observation has one.
    re_block = block_eff[block_id[n]];


    // set both the sequence and population specific effect variables at zero to start, 
    // if it doesn't have a specific effect for
    // the current observation, it will add a 0 deviation to the mu value. 
    re_seq = 0;
    re_pop = 0;

    // if there is a sequence effect, add it now
    if (seq_id[n] > 0) {
      re_seq = seq_eff[seq_id[n]];
    }


    if (pop_rel_id[n] > 0) {
      re_pop = pop_rel_eff[pop_rel_id[n]];
    }

    // If it is an Unreliable population, set the effect 
    if (pop_unrel_id[n] > 0) {
      re_pop = pop_unrel_eff[pop_unrel_id[n]];
    }


    // Calculate the likelhood as a sampling statemnet
    // sum the effects to get the expected mean for observation 
    y[n] ~ normal(mu_fixed + re_block + re_seq + re_pop, sigma_y);
  }
  
  
}

generated quantities {
  // Pointwise log-likelihood values for model comparison
  vector[N] log_lik;

  for (n in 1:N) {
    // Recompute the same linear predictor used in the model block
    real mu_fixed;
    real re_block;
    real re_seq;
    real re_pop;
    real mu;

    // Fixed effects for ancestral or evolved observations
    if (is_evo[n] == 0) {
      mu_fixed = alpha_anc + state_anc[state_id[n]];
    } else {
      int t = evo_type[n];
      mu_fixed =
        (alpha_anc + delta_alpha[t]) +
        (state_anc[state_id[n]] + delta_state[t, state_id[n]]);
    }

    // Block effect
    re_block = block_eff[block_id[n]];

    // set both the sequence and population specific effect variables at zero to start, 
    // if it doesn't have a specific effect for
    // the current observation, it will add a 0 deviation to the mu value. 
    re_seq = 0;
    re_pop = 0;

    // if there is a sequence effect, add it now
    if (seq_id[n] > 0) {
      re_seq = seq_eff[seq_id[n]];
    }


    if (pop_rel_id[n] > 0) {
      re_pop = pop_rel_eff[pop_rel_id[n]];
    }

    // If it is an Unreliable population, set the effect 
    if (pop_unrel_id[n] > 0) {
      re_pop = pop_unrel_eff[pop_unrel_id[n]];
    }

    // sum the effects to get the expected mean for observation 
    mu = mu_fixed + re_block + re_seq + re_pop;

    // Calculate log-likelihood 
    log_lik[n] = normal_lpdf(y[n] | mu, sigma_y);
  }
}

"
```

```{r growth_066_write_stan_full_interaction_only}

stan_file_growth_full_interaction_only <- write_stan_file(
  stan_code_growth_interactionsonly,
  dir = "stan_cache"
)


```



```{r growth_067_compile_stan_full_interaction_only}
mod_growth_full_interaction_only <- cmdstan_model(
  stan_file_growth_full_interaction_only
)
```


```{r growth_068_build_data_full_interaction_only}
stan_data_growth_full_interaction_only <- list(
  N = nrow(d_growth),
  y = as.vector(d_growth$log_growth),
  is_evo = as.integer(d_growth$is_evo),
  evo_type = as.integer(d_growth$evo_type),
  state_id = as.integer(d_growth$state_id),
  B = max(d_growth$block_id),
  block_id = as.integer(d_growth$block_id),
  K_seq = max(d_growth$EE_seq_id),
  seq_id = as.integer(d_growth$EE_seq_id),
  J_rel = max(d_growth$pop_rel_id),
  pop_rel_id = as.integer(d_growth$pop_rel_id),
  J_unrel = max(d_growth$pop_unrel_id),
  pop_unrel_id = as.integer(d_growth$pop_unrel_id)
)

str(stan_data_growth_full_interaction_only, max.level = 1)
```


```{r growth_069_fit_stan_full_interaction_only, cache=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
fit_growth_full_interaction_only <- mod_growth_full_interaction_only$sample(
  data = stan_data_growth_full_interaction_only,
  seed = 42,
  chains = growth_chains,
  parallel_chains = growth_parallel_chains,
  iter_warmup = growth_iter_warmup,
  iter_sampling = growth_iter_sampling,
  adapt_delta = growth_adapt_delta,
  max_treedepth = growth_max_treedepth 
)
```



```{r growth_070_save_stan_full_interaction_only}
 
saveRDS(
  fit_growth_full_interaction_only$draws(),
  "stan_output/growth_full_interaction_only_draws.rds"
)

```





## Setup code to compare models with/without additive and interaction effects

```{r growth_06_write_stan_full_centered}
stan_code_growth_full_centered <- "
functions {
  vector center_sum_to_zero(vector x) {
    return x - mean(x);
  }
}

data {
  int<lower=1> N;
  vector[N] y;

  array[N] int<lower=0, upper=1> is_evo;
  array[N] int<lower=0, upper=2> evo_type;

  array[N] int<lower=-1, upper=1> Ci;
  array[N] int<lower=1, upper=4> mat_id;
  array[N] int<lower=1, upper=8> state_id;

  int<lower=0, upper=1> use_state_anc;
  array[2] int<lower=0, upper=1> use_state;

  int<lower=1> B;
  array[N] int<lower=1, upper=B> block_id;

  int<lower=1> K_seq;
  array[N] int<lower=0, upper=K_seq> seq_id;

  int<lower=1> J_rel;
  array[N] int<lower=0, upper=J_rel> pop_rel_id;

  int<lower=1> J_unrel;
  array[N] int<lower=0, upper=J_unrel> pop_unrel_id;
}

parameters {
  real alpha_anc;

  vector[4] mat_raw_anc;
  real<lower=0> sigma_mat_anc;

  real beta_cue_anc;

  vector[8] state_raw_anc;
  real<lower=0> sigma_state_anc;

  vector[2] delta_alpha;

  matrix[2, 4] delta_mat_raw;
  vector<lower=0>[2] sigma_delta_mat;

  vector[2] delta_beta_cue;

  matrix[2, 8] delta_state_raw;
  vector<lower=0>[2] sigma_delta_state;

  vector[B] z_block;
  real<lower=0> sigma_block;

  vector[K_seq] z_seq;
  real<lower=0> sigma_seq;

  vector[J_rel] z_pop_rel;
  vector[J_unrel] z_pop_unrel;
  real<lower=0> sigma_pop;

  real<lower=0> sigma_y;
}

transformed parameters {
  vector[4] mat_anc   = center_sum_to_zero(sigma_mat_anc * mat_raw_anc);
  vector[8] state_anc = center_sum_to_zero(sigma_state_anc * state_raw_anc);

  matrix[2, 4] delta_mat;
  matrix[2, 8] delta_state;

  for (t in 1:2) {
    delta_mat[t] = (center_sum_to_zero(sigma_delta_mat[t] * to_vector(delta_mat_raw[t])))';
    delta_state[t] = (center_sum_to_zero(sigma_delta_state[t] * to_vector(delta_state_raw[t])))';
  }

  vector[B] block_eff = center_sum_to_zero(sigma_block * z_block);

  vector[K_seq] seq_eff;
  if (K_seq > 0)
    seq_eff = center_sum_to_zero(sigma_seq * z_seq);

  vector[J_rel] pop_rel_eff;
  if (J_rel > 0)
    pop_rel_eff = center_sum_to_zero(sigma_pop * z_pop_rel);

  vector[J_unrel] pop_unrel_eff;
  if (J_unrel > 0)
    pop_unrel_eff = center_sum_to_zero(sigma_pop * z_pop_unrel);
}

model {
  alpha_anc ~ normal(0, 1.5);

  mat_raw_anc ~ normal(0, 1);
  sigma_mat_anc ~ exponential(1);

  beta_cue_anc ~ normal(0, 1);

  state_raw_anc ~ normal(0, 1);
  sigma_state_anc ~ exponential(1);

  delta_alpha ~ normal(0, 1);
  to_vector(delta_mat_raw) ~ normal(0, 1);
  sigma_delta_mat ~ exponential(1);

  delta_beta_cue ~ normal(0, 1);

  to_vector(delta_state_raw) ~ normal(0, 1);
  sigma_delta_state ~ exponential(1);

  z_block ~ normal(0, 1);
  sigma_block ~ exponential(1);

  if (K_seq > 0) {
    z_seq ~ normal(0, 1);
    sigma_seq ~ exponential(1);
  }

  if (J_rel > 0)
    z_pop_rel ~ normal(0, 1);

  if (J_unrel > 0)
    z_pop_unrel ~ normal(0, 1);

  sigma_pop ~ exponential(1);
  sigma_y ~ exponential(1);

  for (n in 1:N) {
    real mu_fixed;

    if (is_evo[n] == 0) {
      if (use_state_anc == 1) {
        mu_fixed = alpha_anc + state_anc[state_id[n]];
      } else {
        mu_fixed = alpha_anc
                 + mat_anc[mat_id[n]]
                 + beta_cue_anc * Ci[n];
      }
    } else {
      int t = evo_type[n];

      if (use_state[t] == 1) {
        mu_fixed = (alpha_anc + delta_alpha[t])
                 + (state_anc[state_id[n]] + delta_state[t, state_id[n]]);
      } else {
        mu_fixed = (alpha_anc + delta_alpha[t])
                 + (mat_anc[mat_id[n]] + delta_mat[t, mat_id[n]])
                 + (beta_cue_anc + delta_beta_cue[t]) * Ci[n];
      }
    }

    {
      real re_block = block_eff[block_id[n]];
      real re_seq = 0;
      real re_pop = 0;

      if (seq_id[n] > 0)
        re_seq = seq_eff[seq_id[n]];

      if (pop_rel_id[n] > 0)
        re_pop += pop_rel_eff[pop_rel_id[n]];

      if (pop_unrel_id[n] > 0)
        re_pop += pop_unrel_eff[pop_unrel_id[n]]; 

      y[n] ~ normal(mu_fixed + re_block + re_seq + re_pop, sigma_y); // likelihood of data evaluation
    }
  }
}

generated quantities {
  vector[N] log_lik;

  for (n in 1:N) {
    real mu_fixed;

    if (is_evo[n] == 0) {
      if (use_state_anc == 1) {
        mu_fixed = alpha_anc + state_anc[state_id[n]];
      } else {
        mu_fixed = alpha_anc
                 + mat_anc[mat_id[n]]
                 + beta_cue_anc * Ci[n];
      }
    } else {
      int t = evo_type[n];

      if (use_state[t] == 1) {
        mu_fixed = (alpha_anc + delta_alpha[t])
                 + (state_anc[state_id[n]] + delta_state[t, state_id[n]]);
      } else {
        mu_fixed = (alpha_anc + delta_alpha[t])
                 + (mat_anc[mat_id[n]] + delta_mat[t, mat_id[n]])
                 + (beta_cue_anc + delta_beta_cue[t]) * Ci[n];
      }
    }

    {
      real re_block = block_eff[block_id[n]];
      real re_seq = 0;
      real re_pop = 0;

      if (seq_id[n] > 0)
        re_seq = seq_eff[seq_id[n]];

      if (pop_rel_id[n] > 0)
        re_pop += pop_rel_eff[pop_rel_id[n]];

      if (pop_unrel_id[n] > 0)
        re_pop += pop_unrel_eff[pop_unrel_id[n]];

      log_lik[n] = normal_lpdf(y[n] | mu_fixed + re_block + re_seq + re_pop, sigma_y);
    }
  }
}
"

stan_file_growth_full_centered <- write_stan_file(stan_code_growth_full_centered, dir = "stan_cache")
```

```{r growth_07_compile_stan_full_centered}
mod_growth_full_centered <- cmdstan_model(stan_file_growth_full_centered)
```


```{r growth_08_fit_switch_grid}
growth_switch_grid <- list(
  growth_additive_all = list(use_state_anc = 0L, use_state = c(0L, 0L)),
  growth_state_all    = list(use_state_anc = 1L, use_state = c(1L, 1L)),
  growth_state_anc    = list(use_state_anc = 1L, use_state = c(0L, 0L)),
  growth_state_rel    = list(use_state_anc = 0L, use_state = c(1L, 0L)),
  growth_state_unrel  = list(use_state_anc = 0L, use_state = c(0L, 1L))
)

fit_growth_switch <- lapply(names(growth_switch_grid), function(nm) {
  spec <- growth_switch_grid[[nm]]
  dat <- stan_data_growth
  dat$use_state_anc <- spec$use_state_anc
  dat$use_state <- as.integer(spec$use_state)

  mod_growth_full_centered$sample(
    data = dat,
  seed = 42,
    chains = growth_chains,
    parallel_chains = growth_parallel_chains,
    iter_warmup = growth_iter_warmup,
    iter_sampling = growth_iter_sampling,
    adapt_delta = growth_adapt_delta,
    max_treedepth = growth_max_treedepth
  )
})
fit_growth_switch <- rlang::set_names(fit_growth_switch, names(growth_switch_grid))
```




```{r growth_09_diagnostics_helpers}
summarize_diag <- function(fit) {
  s <- fit$summary()

  tibble(
    n_params     = nrow(s),
    rhat_max     = max(s$rhat, na.rm = TRUE),
    rhat_gt_1_01 = sum(s$rhat > 1.01, na.rm = TRUE),
    rhat_gt_1_05 = sum(s$rhat > 1.05, na.rm = TRUE),
    ess_bulk_min = min(s$ess_bulk, na.rm = TRUE),
    ess_tail_min = min(s$ess_tail, na.rm = TRUE),
    n_divergent  = sum(fit$diagnostic_summary()$num_divergent, na.rm = TRUE),
    max_treedepth_hits = sum(fit$diagnostic_summary()$num_max_treedepth, na.rm = TRUE)
  ) %>%
    mutate(
      rhat_max = round(rhat_max, 3),
      ess_bulk_min = round(ess_bulk_min, 0),
      ess_tail_min = round(ess_tail_min, 0)
    )
}

summarize_model_comparison <- function(fit_list) {
  loglik_list <- lapply(fit_list, function(fit) {
    posterior::as_draws_matrix(fit$draws("log_lik"))
  })

  loo_list <- lapply(loglik_list, loo::loo)
  stacking_weights <- loo::loo_model_weights(loo_list, method = "stacking")
  loo_table <- as.data.frame(loo::loo_compare(loo_list))
  loo_table$model <- rownames(loo_table)

  loo_table %>%
    as_tibble() %>%
    mutate(weight_stacking = as.numeric(stacking_weights[model])) %>%
    select(model, elpd_diff, se_diff, elpd_loo, p_loo, looic, weight_stacking) %>%
    mutate(across(c(elpd_diff, se_diff, elpd_loo, p_loo, looic, weight_stacking), ~ round(.x, 2)))
}
```

```{r growth_10_run_diagnostics}
growth_diag_table <- purrr::imap_dfr(
  fit_growth_switch,
  ~ summarize_diag(.x) %>% mutate(model = .y),
  .id = NULL
) %>%
  select(model, everything())

growth_switch_table <- summarize_model_comparison(fit_growth_switch)

growth_diag_table
growth_switch_table
```




## Model for the effect of evolutionary sequence
Added April 9, 2026


But actually do it for Unreliable too.

```{r growth_06a_write_stan_anc_rel_unrel_by_seq_state_only}
stan_code_growth_anc_rel_unrel_by_seq_state_only <- "
functions {
  vector center_sum_to_zero(vector x) {
    return x - mean(x);
  }
}

data {
  // Number of observations
  int<lower=1> N;

  // Observed log fitness
  vector[N] y;

  // 0 = Ancestral, 1 = Reliable, 2 = Unreliable
  array[N] int<lower=0, upper=2> evo_type;

  // Full 8-state index
  array[N] int<lower=1, upper=8> state_id;

  // Assay block
  int<lower=1> B;
  array[N] int<lower=1, upper=B> block_id;

  // Number of Reliable sequences
  int<lower=1> K_seq_rel;

  // Number of Unreliable sequences
  int<lower=1> K_seq_unrel;

  // Sequence ids for each observation
  // Use 0 when sequence does not apply
  array[N] int<lower=0, upper=K_seq_rel> seq_id_rel;
  array[N] int<lower=0, upper=K_seq_unrel> seq_id_unrel;

  // Total number of Reliable replicate populations
  int<lower=1> J_rel;

  // Total number of Unreliable replicate populations
  int<lower=1> J_unrel;

  // Population ids for each observation
  // Use 0 when population effect does not apply
  array[N] int<lower=0, upper=J_rel> pop_rel_id;
  array[N] int<lower=0, upper=J_unrel> pop_unrel_id;

  // Sequence membership of each Reliable population
  array[J_rel] int<lower=1, upper=K_seq_rel> pop_seq_of_rel;

  // Sequence membership of each Unreliable population
  array[J_unrel] int<lower=1, upper=K_seq_unrel> pop_seq_of_unrel;
}

parameters {
  // Ancestral mean across states
  real alpha_anc;

  // Ancestral state effects
  vector[8] state_raw_anc;
  real<lower=0> sigma_state_anc;

  // Reliable sequence-specific intercept shifts relative to Ancestral
  vector[K_seq_rel] delta_alpha_rel_seq;

  // Reliable sequence-specific state deviations relative to Ancestral
  matrix[K_seq_rel, 8] delta_state_raw_rel_seq;
  vector<lower=0>[K_seq_rel] sigma_delta_state_rel_seq;

  // Unreliable sequence-specific intercept shifts relative to Ancestral
  vector[K_seq_unrel] delta_alpha_unrel_seq;

  // Unreliable sequence-specific state deviations relative to Ancestral
  matrix[K_seq_unrel, 8] delta_state_raw_unrel_seq;
  vector<lower=0>[K_seq_unrel] sigma_delta_state_unrel_seq;

  // Block effects
  vector[B] z_block;
  real<lower=0> sigma_block;

  // Reliable replicate-population effects
  vector[J_rel] z_pop_rel;
  real<lower=0> sigma_pop_rel;

  // Unreliable replicate-population effects
  vector[J_unrel] z_pop_unrel;
  real<lower=0> sigma_pop_unrel;

  // Residual standard deviation
  real<lower=0> sigma_y;
}

transformed parameters {
  vector[8] state_anc = center_sum_to_zero(sigma_state_anc * state_raw_anc);

  matrix[K_seq_rel, 8] delta_state_rel_seq;
  matrix[K_seq_unrel, 8] delta_state_unrel_seq;

  vector[B] block_eff = center_sum_to_zero(sigma_block * z_block);

  vector[J_rel] pop_rel_eff;
  vector[J_unrel] pop_unrel_eff;

  for (s in 1:K_seq_rel) {
    delta_state_rel_seq[s] =
      to_row_vector(
        center_sum_to_zero(
          sigma_delta_state_rel_seq[s] * to_vector(delta_state_raw_rel_seq[s])
        )
      );
  }

  for (s in 1:K_seq_unrel) {
    delta_state_unrel_seq[s] =
      to_row_vector(
        center_sum_to_zero(
          sigma_delta_state_unrel_seq[s] * to_vector(delta_state_raw_unrel_seq[s])
        )
      );
  }

  // Center Reliable population effects within sequence
  for (s in 1:K_seq_rel) {
    int n_in_seq = 0;
    real mean_z = 0;

    for (j in 1:J_rel) {
      if (pop_seq_of_rel[j] == s) {
        n_in_seq += 1;
        mean_z += z_pop_rel[j];
      }
    }

    mean_z /= n_in_seq;

    for (j in 1:J_rel) {
      if (pop_seq_of_rel[j] == s) {
        pop_rel_eff[j] = sigma_pop_rel * (z_pop_rel[j] - mean_z);
      }
    }
  }

  // Center Unreliable population effects within sequence
  for (s in 1:K_seq_unrel) {
    int n_in_seq = 0;
    real mean_z = 0;

    for (j in 1:J_unrel) {
      if (pop_seq_of_unrel[j] == s) {
        n_in_seq += 1;
        mean_z += z_pop_unrel[j];
      }
    }

    mean_z /= n_in_seq;

    for (j in 1:J_unrel) {
      if (pop_seq_of_unrel[j] == s) {
        pop_unrel_eff[j] = sigma_pop_unrel * (z_pop_unrel[j] - mean_z);
      }
    }
  }
}

model {
  alpha_anc ~ normal(0, 1.5);

  state_raw_anc ~ normal(0, 1);
  sigma_state_anc ~ exponential(1);

  delta_alpha_rel_seq ~ normal(0, 1);
  to_vector(delta_state_raw_rel_seq) ~ normal(0, 1);
  sigma_delta_state_rel_seq ~ exponential(1);

  delta_alpha_unrel_seq ~ normal(0, 1);
  to_vector(delta_state_raw_unrel_seq) ~ normal(0, 1);
  sigma_delta_state_unrel_seq ~ exponential(1);

  z_block ~ normal(0, 1);
  sigma_block ~ exponential(1);

  z_pop_rel ~ normal(0, 1);
  sigma_pop_rel ~ exponential(1);

  z_pop_unrel ~ normal(0, 1);
  sigma_pop_unrel ~ exponential(1);

  sigma_y ~ exponential(1);

  for (n in 1:N) {
    real mu_fixed;
    real re_block;
    real re_pop;

    if (evo_type[n] == 0) {
      mu_fixed =
        alpha_anc +
        state_anc[state_id[n]];
    } else if (evo_type[n] == 1) {
      int s = seq_id_rel[n];
      mu_fixed =
        alpha_anc +
        state_anc[state_id[n]] +
        delta_alpha_rel_seq[s] +
        delta_state_rel_seq[s, state_id[n]];
    } else {
      int s = seq_id_unrel[n];
      mu_fixed =
        alpha_anc +
        state_anc[state_id[n]] +
        delta_alpha_unrel_seq[s] +
        delta_state_unrel_seq[s, state_id[n]];
    }

    re_block = block_eff[block_id[n]];
    re_pop = 0;

    if (pop_rel_id[n] > 0) {
      re_pop = pop_rel_eff[pop_rel_id[n]];
    }
    if (pop_unrel_id[n] > 0) {
      re_pop = pop_unrel_eff[pop_unrel_id[n]];
    }

    y[n] ~ normal(mu_fixed + re_block + re_pop, sigma_y);
  }
}

generated quantities {
  vector[N] log_lik;
  vector[8] mu_anc_state;
  matrix[K_seq_rel, 8] mu_rel_state;
  matrix[K_seq_unrel, 8] mu_unrel_state;

  for (st in 1:8) {
    mu_anc_state[st] = alpha_anc + state_anc[st];
  }

  for (s in 1:K_seq_rel) {
    for (st in 1:8) {
      mu_rel_state[s, st] =
        alpha_anc +
        state_anc[st] +
        delta_alpha_rel_seq[s] +
        delta_state_rel_seq[s, st];
    }
  }

  for (s in 1:K_seq_unrel) {
    for (st in 1:8) {
      mu_unrel_state[s, st] =
        alpha_anc +
        state_anc[st] +
        delta_alpha_unrel_seq[s] +
        delta_state_unrel_seq[s, st];
    }
  }

  for (n in 1:N) {
    real mu_fixed;
    real re_block;
    real re_pop;
    real mu;

    if (evo_type[n] == 0) {
      mu_fixed =
        alpha_anc +
        state_anc[state_id[n]];
    } else if (evo_type[n] == 1) {
      int s = seq_id_rel[n];
      mu_fixed =
        alpha_anc +
        state_anc[state_id[n]] +
        delta_alpha_rel_seq[s] +
        delta_state_rel_seq[s, state_id[n]];
    } else {
      int s = seq_id_unrel[n];
      mu_fixed =
        alpha_anc +
        state_anc[state_id[n]] +
        delta_alpha_unrel_seq[s] +
        delta_state_unrel_seq[s, state_id[n]];
    }

    re_block = block_eff[block_id[n]];
    re_pop = 0;

    if (pop_rel_id[n] > 0) {
      re_pop = pop_rel_eff[pop_rel_id[n]];
    }
    if (pop_unrel_id[n] > 0) {
      re_pop = pop_unrel_eff[pop_unrel_id[n]];
    }

    mu = mu_fixed + re_block + re_pop;
    log_lik[n] = normal_lpdf(y[n] | mu, sigma_y);
  }
}
"
```


```{r growth_06b_build_data_anc_rel_unrel_by_seq_state_only}
d_growth_anc_rel_unrel_seq <- d_growth %>%
  dplyr::filter(evo_type %in% c(0L, 1L, 2L))

# Re-index Reliable sequences so they run 1:K_seq_rel within this subset
rel_seq_levels <- d_growth_anc_rel_unrel_seq %>%
  dplyr::filter(evo_type == 1L) %>%
  dplyr::distinct(EE_seq_id) %>%
  dplyr::arrange(EE_seq_id) %>%
  dplyr::pull(EE_seq_id)

# Re-index Unreliable sequences so they run 1:K_seq_unrel within this subset
unrel_seq_levels <- d_growth_anc_rel_unrel_seq %>%
  dplyr::filter(evo_type == 2L) %>%
  dplyr::distinct(EE_seq_id) %>%
  dplyr::arrange(EE_seq_id) %>%
  dplyr::pull(EE_seq_id)

d_growth_anc_rel_unrel_seq <- d_growth_anc_rel_unrel_seq %>%
  dplyr::mutate(
    seq_id_rel = dplyr::if_else(
      evo_type == 1L,
      match(EE_seq_id, rel_seq_levels),
      0L
    ),
    seq_id_unrel = dplyr::if_else(
      evo_type == 2L,
      match(EE_seq_id, unrel_seq_levels),
      0L
    )
  )

# Re-index Reliable populations within this subset
rel_pop_levels_seq <- d_growth_anc_rel_unrel_seq %>%
  dplyr::filter(evo_type == 1L) %>%
  dplyr::distinct(population) %>%
  dplyr::arrange(population) %>%
  dplyr::pull(population)

# Re-index Unreliable populations within this subset
unrel_pop_levels_seq <- d_growth_anc_rel_unrel_seq %>%
  dplyr::filter(evo_type == 2L) %>%
  dplyr::distinct(population) %>%
  dplyr::arrange(population) %>%
  dplyr::pull(population)

d_growth_anc_rel_unrel_seq <- d_growth_anc_rel_unrel_seq %>%
  dplyr::mutate(
    pop_rel_id = dplyr::if_else(
      evo_type == 1L,
      match(population, rel_pop_levels_seq),
      0L
    ),
    pop_unrel_id = dplyr::if_else(
      evo_type == 2L,
      match(population, unrel_pop_levels_seq),
      0L
    )
  )

# Map each Reliable population to its sequence
pop_seq_of_rel <- d_growth_anc_rel_unrel_seq %>%
  dplyr::filter(evo_type == 1L) %>%
  dplyr::distinct(pop_rel_id, seq_id_rel) %>%
  dplyr::arrange(pop_rel_id) %>%
  dplyr::pull(seq_id_rel)

# Map each Unreliable population to its sequence
pop_seq_of_unrel <- d_growth_anc_rel_unrel_seq %>%
  dplyr::filter(evo_type == 2L) %>%
  dplyr::distinct(pop_unrel_id, seq_id_unrel) %>%
  dplyr::arrange(pop_unrel_id) %>%
  dplyr::pull(seq_id_unrel)

stan_data_growth_anc_rel_unrel_by_seq_state_only <- list(
  N = nrow(d_growth_anc_rel_unrel_seq),
  y = as.vector(d_growth_anc_rel_unrel_seq$log_growth),
  evo_type = as.integer(d_growth_anc_rel_unrel_seq$evo_type),
  state_id = as.integer(d_growth_anc_rel_unrel_seq$state_id),
  B = max(d_growth_anc_rel_unrel_seq$block_id),
  block_id = as.integer(d_growth_anc_rel_unrel_seq$block_id),
  K_seq_rel = length(rel_seq_levels),
  K_seq_unrel = length(unrel_seq_levels),
  seq_id_rel = as.integer(d_growth_anc_rel_unrel_seq$seq_id_rel),
  seq_id_unrel = as.integer(d_growth_anc_rel_unrel_seq$seq_id_unrel),
  J_rel = length(rel_pop_levels_seq),
  J_unrel = length(unrel_pop_levels_seq),
  pop_rel_id = as.integer(d_growth_anc_rel_unrel_seq$pop_rel_id),
  pop_unrel_id = as.integer(d_growth_anc_rel_unrel_seq$pop_unrel_id),
  pop_seq_of_rel = as.integer(pop_seq_of_rel),
  pop_seq_of_unrel = as.integer(pop_seq_of_unrel)
)

str(stan_data_growth_anc_rel_unrel_by_seq_state_only, max.level = 1)

```




```{r growth_06c_compile_stan_anc_rel_unrel_by_seq_state_only}
stan_file_growth_anc_rel_unrel_by_seq_state_only <- write_stan_file(
  stan_code_growth_anc_rel_unrel_by_seq_state_only,
  dir = "stan_cache"
)

mod_growth_anc_rel_unrel_by_seq_state_only <- cmdstan_model(
  stan_file_growth_anc_rel_unrel_by_seq_state_only
)
```



```{r growth_06d_fit_stan_anc_rel_unrel_by_seq_state_only, cache=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
fit_growth_anc_rel_unrel_by_seq_state_only <- mod_growth_anc_rel_unrel_by_seq_state_only$sample(
  data = stan_data_growth_anc_rel_unrel_by_seq_state_only,
  seed = 42,
  chains = growth_chains,
  parallel_chains = growth_parallel_chains,
  iter_warmup = growth_iter_warmup,
  iter_sampling = growth_iter_sampling,
  adapt_delta = growth_adapt_delta,
  max_treedepth = growth_max_treedepth
)
```


```{r growth_06e_save_stan_anc_rel_unrel_by_seq_state_only}


saveRDS(
  fit_growth_anc_rel_unrel_by_seq_state_only$draws(),
  "stan_output/growth_anc_rel_unrel_by_seq_state_only_draws.rds"
)

```





