---
title: "Combined_Growth_and_Fecundity_Plots"
author: "Stephen R. Proulx"
date: "2026-03-20"
output:
  pdf_document:
    keep_tex: true
    latex_engine: xelatex
---


```{r growth_plot_helpers_minimal, include=FALSE}
library(tidyverse)
library(posterior)
library(grid)

draws_growth_state_all <- readRDS("stan_output/growth_full_interaction_only_draws.rds")
growth_draws_df <- posterior::as_draws_df(draws_growth_state_all)

growth_key_mat_levels <- c("AN", "NN", "AA", "NA")
growth_cue_plot_levels <- c("No Light", "Blue Light")



draws_fec_state_all <- readRDS("stan_output/fit_fecundity_additive_draws_df.rds")
fec_draws_df <- posterior::as_draws_df(draws_fec_state_all)



growth_cue_plot_colors <- c(
  "No Light" = "#999999",
  "Blue Light" = "#0072B2"
)

fec_cue_plot_colors <- c(
  "No Light" = "#999999",
  "Blue Light" = "#0072B2"
)


contrast_color <- "#7A3E9D"

growth_base_plot_theme <- function(base_size = 16) {
  theme_classic(base_size = base_size) +
    theme(
      legend.position = "right",
      strip.background = element_rect(
        fill = "white",
        color = "black",
        linewidth = 0.8
      ),
      strip.text = element_text(
        face = "bold",
        size = 12,
        margin = margin(4, 6, 4, 6)
      )
    )
}

fec_base_plot_theme <- function(base_size = 16) {
  theme_classic(base_size = base_size) +
    theme(
      legend.position = "right",
      strip.background = element_rect(
        fill = "white",
        color = "black",
        linewidth = 0.8
      ),
      strip.text = element_text(
        face = "bold",
        size = 12,
        margin = margin(4, 6, 4, 6)
      )
    )
}


hide_x_axis_text_theme <- function() {
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
}

axis_label_growth <- "Fitness"

axis_label_fecundity <- "Fecundity"

axis_label_mat_env_full <- "Oxygen Environment"
legend_label_light_cue <- "Light Environment"

growth_lo_p <- 0.055
growth_hi_p <- 0.945

fec_lo_p <- 0.055
fec_hi_p <- 0.945
```


re-usable features for growth data



```{r constants , include=FALSE}

mat_key_growth <- tibble::tibble(
  mat_id = 1:4,
  mat_label = c("AN", "NN", "AA", "NA")
)

state_key_growth <- tibble::tibble(
  state_id = 1:8,
  mat_id = c(1, 1, 2, 2, 3, 3, 4, 4),
  mat_label = c("AN", "AN", "NN", "NN", "AA", "AA", "NA", "NA"),
  cue = c("No Light", "Blue Light", "No Light", "Blue Light",
          "No Light", "Blue Light", "No Light", "Blue Light")
) 

# ---- Fecundity keys: match 02_run_Fecundity_additive.R exactly ----
mat_key_fec <- tibble::tibble(
  mat_id = 1:4,
  mat = c("AA", "AN", "NA", "NN")
)

state_key_fec <- tidyr::expand_grid(
  mat = c("AA", "AN", "NA", "NN"),
  cue = c("No Light", "Blue Light")
) %>%
  dplyr::mutate(state_id = dplyr::row_number()) %>%
  dplyr::left_join(mat_key_fec, by = "mat") %>%
  dplyr::select(state_id, mat_id, mat, cue)


plot_mat_levels <- c("AN", "NN", "AA", "NA")
```


plot functions

```{r add_functions , include=FALSE}



growth_plot_halfeye_with_summary <- function(
  df_draws,
  df_summ,
  x,
  y,
  fill = NULL,
  group = NULL,
  x_label = NULL,
  y_label,
  hline_value = 0,
  show_legend = FALSE,
  remove_x_text = FALSE,
  point_size = 2.3,
  hline_width = 0.7,
  dodge_width = NULL,
  single_fill = contrast_color,
  fill_values = growth_cue_plot_colors
) {
  dodge_position <- if (!is.null(dodge_width)) position_dodge(width = dodge_width) else "identity"

  if (!is.null(fill)) {
    p <- ggplot() +
      geom_hline(yintercept = hline_value, linewidth = hline_width) +
      ggdist::stat_halfeye(
        data = df_draws,
        aes(
          x = .data[[x]],
          y = .data[[y]],
          fill = .data[[fill]],
          group = .data[[group]]
        ),
        position = dodge_position,
        side = "right",
        justification = -0.15,
        width = 0.34,
        point_interval = NULL,
        alpha = 0.8,
        slab_color = NA,
        show.legend = show_legend
      ) +
      geom_errorbar(
        data = df_summ,
        aes(
          x = .data[[x]],
          y = .data[["med"]],
          ymin = .data[["lo"]],
          ymax = .data[["hi"]],
          group = .data[[group]]
        ),
        position = dodge_position,
        width = 0,
        linewidth = 0.7,
        color = "black"
      ) +
      geom_point(
        data = df_summ,
        aes(
          x = .data[[x]],
          y = .data[["med"]],
          group = .data[[group]]
        ),
        position = dodge_position,
        size = point_size,
        color = "black"
      ) +
      scale_fill_manual(name = legend_label_light_cue, values = fill_values)
  } else {
    p <- ggplot() +
      geom_hline(yintercept = hline_value, linewidth = hline_width) +
      ggdist::stat_halfeye(
        data = df_draws,
        aes(x = .data[[x]], y = .data[[y]]),
        side = "right",
        justification = -0.15,
        width = 0.34,
        point_interval = NULL,
        alpha = 0.8,
        fill = single_fill,
        slab_color = NA
      ) +
      geom_errorbar(
        data = df_summ,
        aes(
          x = .data[[x]],
          y = .data[["med"]],
          ymin = .data[["lo"]],
          ymax = .data[["hi"]]
        ),
        width = 0,
        linewidth = 0.7,
        color = "black"
      ) +
      geom_point(
        data = df_summ,
        aes(x = .data[[x]], y = .data[["med"]]),
        size = point_size,
        color = "black"
      )
  }

  p <- p +
    labs(x = x_label, y = y_label) +
    growth_base_plot_theme()

  if (remove_x_text) {
    p <- p + hide_x_axis_text_theme()
  }

  p
}

fill_legend_no_point <- function() {
  guides(fill = guide_legend(override.aes = list(shape = NA, color = NA)))
}

boxed_strip_theme <- function(base_size = 12) {
  theme(
    strip.background = element_rect(
      fill = "white",
      color = "black",
      linewidth = 0.8
    ),
    strip.text = element_text(
      face = "bold",
      size = base_size,
      margin = margin(4, 6, 4, 6)
    )
  )
}


```



## Fitness data 

```{r growth_plot_anc_growth_by_matcue_intervals_from_state_only, echo=FALSE, message=FALSE, warning=FALSE}
alpha_fix <- median(growth_draws_df$alpha_anc)

state_df <- growth_draws_df %>%
  select(starts_with("state_anc[")) %>%
  mutate(across(everything(), ~ exp(alpha_fix + .)))

plot_summ <- state_df %>%
  pivot_longer(
    cols = starts_with("state_anc["),
    names_to = "state",
    values_to = "fitness"
  ) %>%
  mutate(
    state_id = as.integer(substr(state, 11, 11)),
    mat_id = (state_id + 1L) %/% 2L)%>%
  left_join(state_key_growth, by = "state_id") %>%
   mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  ) %>%
  group_by(mat_label, cue) %>%
  summarise(
    med = median(fitness),
    lo = quantile(fitness, growth_lo_p),
    hi = quantile(fitness, growth_hi_p),
    .groups = "drop"
  )

ggplot(plot_summ, aes(x = mat_label, y = med, fill = cue)) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.8
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.9),
    width = 0.2,
    linewidth = 0.7
  ) +
  scale_fill_manual(
    name = legend_label_light_cue,
    values = growth_cue_plot_colors
  ) +
  labs(
    x = axis_label_mat_env_full,
    y = axis_label_growth
  ) +
  growth_base_plot_theme()
```


```{r growth_plot_anc_rel_unrel_growth_by_matcue_intervals_state_only, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4.5}
alpha_fix_anc <- median(growth_draws_df$alpha_anc)
alpha_fix_rel <- median(growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[1]`)
alpha_fix_unrel <- median(growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[2]`)

state_df_anc <- growth_draws_df %>%
  select(starts_with("state_anc[")) %>%
  mutate(across(everything(), ~ exp(alpha_fix_anc + .)))  

names(state_df_anc) <- as.character(1:8)

state_df_rel <- growth_draws_df %>%
  select(starts_with("state_anc["), starts_with("delta_state[1,"))

state_df_rel_sum <- state_df_rel[1:8] + state_df_rel[9:16]

names(state_df_rel_sum) <- as.character(1:8)

state_df_rel_sum <- state_df_rel_sum %>%
  as_tibble() %>%
  mutate(across(everything(), ~ exp(alpha_fix_rel + .)))

state_df_unrel <- growth_draws_df %>%
  select(starts_with("state_anc["), starts_with("delta_state[2,"))

state_df_unrel_sum <- state_df_unrel[1:8] + state_df_unrel[9:16]

names(state_df_unrel_sum) <- as.character(1:8)

state_df_unrel_sum <- state_df_unrel_sum %>%
  as_tibble() %>%
  mutate(across(everything(), ~ exp(alpha_fix_unrel + .)))

plot_anc <- state_df_anc %>%
  pivot_longer(
    cols = everything(),
    names_to = "state",
    values_to = "grow"
  ) %>%
  mutate(
    state_id = as.integer(state),
    treatment = "Ancestral"
  ) %>%
  left_join(state_key_growth, by = "state_id")%>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

plot_rel <- state_df_rel_sum %>%
  pivot_longer(
    cols = everything(),
    names_to = "state",
    values_to = "grow"
  ) %>%
  mutate(
    state_id = as.integer(state),
    treatment = "Reliable"
  ) %>%
  left_join(state_key_growth, by = "state_id")%>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

plot_unrel <- state_df_unrel_sum %>%
  pivot_longer(
    cols = everything(),
    names_to = "state",
    values_to = "grow"
  ) %>%
  mutate(
    state_id = as.integer(state),
    treatment = "Unreliable"
  ) %>%
  left_join(state_key_growth, by = "state_id")%>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

plot_summ <- bind_rows(plot_anc, plot_rel, plot_unrel) %>%
  group_by(treatment, mat_label, cue) %>%
  summarise(
    med = median(grow),
    lo = quantile(grow, growth_lo_p),
    hi = quantile(grow, growth_hi_p),
    .groups = "drop"
  )

ggplot(plot_summ, aes(x = mat_label, y = med, fill = cue)) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.8
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.9),
    width = 0.2,
    linewidth = 0.7
  ) +
  scale_fill_manual(
    name = legend_label_light_cue,
    values = growth_cue_plot_colors
  ) +
  labs(
    x = axis_label_mat_env_full,
    y = axis_label_growth
  ) +
  facet_wrap(~ treatment) +
  growth_base_plot_theme()

```


This figure shows the difference in R0 from the ancestral for reliable and unreliable. i.e. R0_rel-R0_anc and R0_unrel-R0_anc. So the pop growth parameters are calculated, exp'ed, and then difference taken.


```{r growth_plot_rel_unrel_minus_anc_growth_contrast_boxplot_style_fecundity_like, echo=FALSE, message=FALSE, warning=FALSE}
alpha_anc <- growth_draws_df$alpha_anc # get the intercept for anc
alpha_rel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[1]` # get the intercept for anc Plus the difference for reliable
alpha_unrel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[2]` # get the intercept for anc Plus the difference for unreliable

#calc anc R0 from median intercept
state_df_anc <- growth_draws_df %>%
  select(starts_with("state_anc[")) %>%
  mutate(across(everything(), ~ exp(alpha_anc + .))) #add intercept and exp()

names(state_df_anc) <- as.character(1:8) #convert column names to state number


#calc Rel R0 from median intercept
state_df_rel <- growth_draws_df %>%
  select(starts_with("state_anc["), starts_with("delta_state[1,")) #anc state 1-8, del rel 9-16

state_df_rel_sum <- (state_df_rel[1:8] + state_df_rel[9:16] + alpha_rel) %>% #add anc +delRel + intercept
  as_tibble() %>%
  mutate(across(everything(), exp)) #convert to natural scale of R0
names(state_df_rel_sum) <- as.character(1:8)


#calc Unrel R0 from median intercept
state_df_unrel <- growth_draws_df %>%
  select(starts_with("state_anc["), starts_with("delta_state[1,")) #anc state 1-8, del unrel 9-16

state_df_unrel_sum <- (state_df_unrel[1:8] + state_df_unrel[9:16] + alpha_unrel) %>% #add anc +delRel + intercept
  as_tibble() %>%
  mutate(across(everything(), exp)) #convert to natural scale of R0
names(state_df_unrel_sum) <- as.character(1:8)


# make anc info into a long form tidy df and join to state_key_growth to get names of Mat env and cue
draws_anc <- state_df_anc %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_anc"
  ) %>% 
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id")%>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

# make rel info into a long form tidy df and join to state_key_growth to get names of Mat env and cue
draws_rel <- state_df_rel_sum %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_rel"
  ) %>%
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id")%>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

# make unrel info into a long form tidy df and join to state_key_growth to get names of Mat env and cue
draws_unrel <- state_df_unrel_sum %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_unrel"
  ) %>%
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id")%>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

#put the Anc and Rel  into a single df and calc the difference in R0
df_draws_rel <- draws_rel %>%
  select(.draw, state_id, mat_label, cue, R0_rel) %>%
  left_join(
    draws_anc %>% select(.draw, state_id, R0_anc),
    by = c(".draw", "state_id")
  ) %>%
  mutate(
    diff_fit = R0_rel - R0_anc,
    contrast = factor(
      "Reliable - Ancestral",
      levels = c("Reliable - Ancestral", "Unreliable - Ancestral")
    )
  )

#put the Anc and Unrel  into a single df and calc the difference in R0
df_draws_unrel <- draws_unrel %>%
  select(.draw, state_id, mat_label, cue, R0_unrel) %>%
  left_join(
    draws_anc %>% select(.draw, state_id, R0_anc),
    by = c(".draw", "state_id")
  ) %>%
  mutate(
    diff_fit = R0_unrel - R0_anc,
    contrast = factor(
      "Unreliable - Ancestral",
      levels = c("Reliable - Ancestral", "Unreliable - Ancestral")
    )
  )


# combine Rel and Unrel dfs
df_draws_both <- bind_rows(df_draws_rel, df_draws_unrel)

#summarise median and quantiles by 
df_summ_both <- df_draws_both %>%
  group_by(contrast, mat_label, cue) %>%
  summarise(
    med = median(diff_fit),
    lo = quantile(diff_fit, growth_lo_p),
    hi = quantile(diff_fit, growth_hi_p),
    .groups = "drop"
  )


#display plot
growth_plot_halfeye_with_summary(
  df_draws = df_draws_both,
  df_summ = df_summ_both,
  x = "mat_label",
  y = "diff_fit",
  fill = "cue",
  group = "cue",
  x_label = axis_label_mat_env_full,
  y_label = "Fitness Divergence",
  show_legend = TRUE,
  dodge_width = 0.7
) +
  facet_wrap(~ contrast, nrow = 1) +
  fill_legend_no_point() +
  boxed_strip_theme(12) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )+
  ylim(-2,2)
```


Plot the cue effect. This finds the R0 values for each state for each treatment, then subtracts Rel-Anc or Unrel-Anc, so outcome is difference in R0

```{r growth_plot_rel_unrel_vs_anc_cue_effect_contrast_boxplot_style_combined, echo=FALSE, message=FALSE, warning=FALSE}
alpha_anc <- growth_draws_df$alpha_anc
alpha_rel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[1]`
alpha_unrel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[2]`

# Calculate ancestral R0 for each state.
# Add the ancestral median intercept then exp
state_df_anc <- growth_draws_df %>%
  select(starts_with("state_anc[")) %>%
  mutate(across(everything(), ~ exp(alpha_anc + .)))

names(state_df_anc) <- as.character(1:8) #convert column names to state number

# Calculate Reliable R0 by state.
#   ancestral state effect + Reliable deviation from ancestral + Reliable intercept
#  then exp to return to the natural scale.
state_df_rel <- growth_draws_df %>%
  select(starts_with("state_anc["), starts_with("delta_state[1,"))

state_df_rel_sum <- (state_df_rel[1:8] + state_df_rel[9:16] + alpha_rel) %>%
  as_tibble() %>%
  mutate(across(everything(), exp))

names(state_df_rel_sum) <- as.character(1:8)


# Calculate Unreliable R0 by state.
#   ancestral state effect + Unreliable deviation from ancestral + Unreliable intercept
#  then exp to return to the natural scale.
state_df_unrel <- growth_draws_df %>%
  select(starts_with("state_anc["), starts_with("delta_state[2,"))

state_df_unrel_sum <- (state_df_unrel[1:8] + state_df_unrel[9:16] + alpha_unrel) %>%
  as_tibble() %>%
  mutate(across(everything(), exp))

names(state_df_unrel_sum) <- as.character(1:8)

# Convert ancestral predictions to long format and attach the state key
# so each draw is labeled by maternal environment and cue.
draws_anc <- state_df_anc %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_anc"
  ) %>%
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id")%>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

# Do the same for Reliable predictions.
draws_rel <- state_df_rel_sum %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_rel"
  ) %>%
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id")%>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

# Do the same for Unreliable predictions.
draws_unrel <- state_df_unrel_sum %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_unrel"
  ) %>%
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id")%>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

# Within each posterior draw and maternal environment, compute the cue effect
# for the ancestral population as a difference on the natural scale:
#   R0 cue - R0  No cue
cue_effect_anc <- draws_anc %>%
  select(.draw, mat_label, cue, R0_anc) %>%
  pivot_wider(
    names_from = cue,
    values_from = R0_anc
  ) %>%
  mutate(
    cue_effect_anc = `Blue Light` - `No Light`
  ) %>%
  select(.draw, mat_label, cue_effect_anc)

# Compute the analogous cue effect for the Reliable population.
cue_effect_rel <- draws_rel %>%
  select(.draw, mat_label, cue, R0_rel) %>%
  pivot_wider(
    names_from = cue,
    values_from = R0_rel
  ) %>%
  mutate(
    cue_effect_rel = `Blue Light` - `No Light`
  ) %>%
  select(.draw, mat_label, cue_effect_rel)

# Compute the analogous cue effect for the Unreliable population.
cue_effect_unrel <- draws_unrel %>%
  select(.draw, mat_label, cue, R0_unrel) %>%
  pivot_wider(
    names_from = cue,
    values_from = R0_unrel
  ) %>%
  mutate(
    cue_effect_unrel = `Blue Light` - `No Light`
  ) %>%
  select(.draw, mat_label, cue_effect_unrel)

# Compare Reliable to Ancestral within each posterior draw and maternal
# environment. A value of 0 means no change in cue response relative to
# the ancestor.
df_draws_rel <- cue_effect_rel %>%
  left_join(
    cue_effect_anc,
    by = c(".draw", "mat_label")
  ) %>%
  mutate(
    cue_effect_diff = cue_effect_rel - cue_effect_anc,
    contrast = factor(
      "Reliable - Ancestral",
      levels = c("Reliable - Ancestral", "Unreliable - Ancestral")
    )
  ) %>%
  select(.draw, mat_label, cue_effect_diff, contrast)

# Compare Unreliable to Ancestral in the same way.
df_draws_unrel <- cue_effect_unrel %>%
  left_join(
    cue_effect_anc,
    by = c(".draw", "mat_label")
  ) %>%
  mutate(
    cue_effect_diff = cue_effect_unrel - cue_effect_anc,
    contrast = factor(
      "Unreliable - Ancestral",
      levels = c("Reliable - Ancestral", "Unreliable - Ancestral")
    )
  ) %>%
  select(.draw, mat_label, cue_effect_diff, contrast)

# Combine the two treatment-vs-ancestor contrasts into one data frame
# so they can be plotted side by side with matching scales.
df_draws_both <- bind_rows(df_draws_rel, df_draws_unrel)

# Summarize the posterior distribution within each panel and maternal
# environment using the median and central posterior interval.
df_summ_both <- df_draws_both %>%
  group_by(contrast, mat_label) %>%
  summarise(
    med = median(cue_effect_diff),
    lo = quantile(cue_effect_diff, growth_lo_p),
    hi = quantile(cue_effect_diff, growth_hi_p),
    .groups = "drop"
  )

# Plot the posterior distributions with median points and interval bars.
# The horizontal reference line at 0 indicates no evolutionary change
# in the cue effect relative to the ancestor.
growth_plot_halfeye_with_summary(
  df_draws = df_draws_both,
  df_summ = df_summ_both,
  x = "mat_label",
  y = "cue_effect_diff",
  x_label = axis_label_mat_env_full,
  y_label = "Light-Plasticity Fitness Divergence",
  hline_value = 0,
  single_fill = unname(contrast_color)
) +
  facet_wrap(~ contrast, nrow = 1) +
  boxed_strip_theme(12)

```



this is the difference of the difference between Rel and Unrel. First calculate the difference in R0 based on cue for Rel, and Unrel. Then do Rel-Unrel of that difference. 

```{r growth_plot_rel_vs_unrel_cue_effect_difference_contrast_boxplot_style, echo=FALSE, message=FALSE, warning=FALSE}
alpha_rel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[1]`
alpha_unrel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[2]`

# Calculate Reliable R0 for each of the 8 maternal-environment x cue states.
state_df_rel <- growth_draws_df %>%
  select(starts_with("state_anc["), starts_with("delta_state[1,"))

state_df_rel_sum <- (state_df_rel[1:8] + state_df_rel[9:16] + alpha_rel) %>%
  as_tibble() %>%
  mutate(across(everything(), exp))

names(state_df_rel_sum) <- as.character(1:8)

# Calculate Unreliable R0 for each of the 8 maternal-environment x cue states.
state_df_unrel <- growth_draws_df %>%
  select(starts_with("state_anc["), starts_with("delta_state[2,"))

state_df_unrel_sum <- (state_df_unrel[1:8] + state_df_unrel[9:16] + alpha_unrel) %>%
  as_tibble() %>%
  mutate(across(everything(), exp))

names(state_df_unrel_sum) <- as.character(1:8)

# Convert Reliable predictions to long format.
draws_rel <- state_df_rel_sum %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_rel"
  ) %>%
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id") %>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

# Convert Unreliable predictions to long format.
draws_unrel <- state_df_unrel_sum %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_unrel"
  ) %>%
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id") %>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

# Reliable cue effect within each posterior draw and maternal environment.
cue_effect_rel <- draws_rel %>%
  select(.draw, mat_label, cue, R0_rel) %>%
  pivot_wider(
    names_from = cue,
    values_from = R0_rel
  ) %>%
  mutate(
    cue_effect_rel = `Blue Light` - `No Light`
  ) %>%
  select(.draw, mat_label, cue_effect_rel)

# Unreliable cue effect within each posterior draw and maternal environment.
cue_effect_unrel <- draws_unrel %>%
  select(.draw, mat_label, cue, R0_unrel) %>%
  pivot_wider(
    names_from = cue,
    values_from = R0_unrel
  ) %>%
  mutate(
    cue_effect_unrel = `Blue Light` - `No Light`
  ) %>%
  select(.draw, mat_label, cue_effect_unrel)

# Maternal-environment-specific Reliable - Unreliable cue-effect differences.
df_draws_mat <- cue_effect_rel %>%
  left_join(
    cue_effect_unrel,
    by = c(".draw", "mat_label")
  ) %>%
  mutate(
    cue_effect_diff = cue_effect_rel - cue_effect_unrel,
    plot_label = as.character(mat_label)
  )

# Overall average cue-effect difference.
# This averages the four maternal-environment-specific differences within
# each posterior draw, so uncertainty is preserved across posterior draws.
df_draws_overall <- df_draws_mat %>%
  group_by(.draw) %>%
  summarise(
    cue_effect_rel = mean(cue_effect_rel),
    cue_effect_unrel = mean(cue_effect_unrel),
    cue_effect_diff = mean(cue_effect_diff),
    plot_label = "Overall",
    .groups = "drop"
  )

# Combine the four maternal-environment distributions with the overall average.
df_draws <- bind_rows(
  df_draws_mat,
  df_draws_overall
)

# Use numeric x positions so the Overall distribution can be spatially separated.
plot_x_positions <- c(
  "AN" = 1,
  "NN" = 2,
  "AA" = 3,
  "NA" = 4,
  "Overall" = 5.75
)

df_draws <- df_draws %>%
  mutate(
    x_pos = unname(plot_x_positions[plot_label])
  )

# Summarize each posterior distribution using the median and central interval.
df_summ <- df_draws %>%
  group_by(plot_label, x_pos) %>%
  summarise(
    med = median(cue_effect_diff),
    lo = quantile(cue_effect_diff, growth_lo_p),
    hi = quantile(cue_effect_diff, growth_hi_p),
    .groups = "drop"
  )

# Plot the four maternal-environment contrasts plus the overall average.
growth_plot_halfeye_with_summary(
  df_draws = df_draws,
  df_summ = df_summ,
  x = "x_pos",
  y = "cue_effect_diff",
  x_label = axis_label_mat_env_full,
  y_label = "Light-Plasticity Fitness Differentiation",
  hline_value = 0,
  single_fill = unname(contrast_color)
) +
  geom_vline(
    xintercept = 4.875,
    linetype = "dotted",
    linewidth = 0.5
  ) +
  scale_x_continuous(
    breaks = unname(plot_x_positions),
    labels = names(plot_x_positions),
    limits = c(0.5, 6.25)
  )
```


Here the contrasts in mean log fitness are calculated on log scale (because it is geo mean of R0). We then plot fixed-reference changes on the R0 scale so they can be seen in terms of number of offspring. We do this by taking exp( Anc  + 1/2 diff in log fit due to evo treat)-exp( Anc  - 1/2 diff in log fit due to evo treat)


```{r geo_mean , echo=FALSE, message=FALSE, warning=FALSE}

alpha_anc <- growth_draws_df$alpha_anc
alpha_rel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[1]`
alpha_unrel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[2]`


# STEP 1: Build log-scale state predictions for each treatment


state_df_anc_log <- growth_draws_df %>%
  select(paste0("state_anc[", 1:8, "]")) %>%
  mutate(across(everything(), ~ alpha_anc + .))

names(state_df_anc_log) <- as.character(1:8)

state_df_rel <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[1,", 1:8, "]")
  )

state_df_rel_log <- (state_df_rel[1:8] + state_df_rel[9:16] + alpha_rel) %>%
  as_tibble()

names(state_df_rel_log) <- as.character(1:8)

state_df_unrel <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[2,", 1:8, "]")
  )

state_df_unrel_log <- (state_df_unrel[1:8] + state_df_unrel[9:16] + alpha_unrel) %>%
  as_tibble()

names(state_df_unrel_log) <- as.character(1:8)


# STEP 2: Average on the log scale
# In this file, Blue Light states are 2, 4, 6, 8


blue_state_ids <- c(2, 4, 6, 8)
all_state_ids <- 1:8

avg_log_anc_blue <- rowMeans(state_df_anc_log[, blue_state_ids])
avg_log_rel_blue <- rowMeans(state_df_rel_log[, blue_state_ids])

avg_log_anc_all <- rowMeans(state_df_anc_log[, all_state_ids])
avg_log_unrel_all <- rowMeans(state_df_unrel_log[, all_state_ids])


# STEP 3: Log-scale contrasts
# Match the existing plotting logic exactly


diff_rel_vs_anc <- avg_log_rel_blue - avg_log_anc_blue
diff_unrel_vs_anc <- avg_log_unrel_all - avg_log_anc_all


# STEP 4: Fixed ancestral reference on the log scale
# Use the median overall Anc value on the log scale


median_anc_log <- median(avg_log_anc_all)


# STEP 5: Map the log-scale contrast back to the natural scale
# using the symmetric transformation around the fixed reference


diff_rel_nat <- exp(median_anc_log + 0.5 * diff_rel_vs_anc) -
  exp(median_anc_log - 0.5 * diff_rel_vs_anc)

diff_unrel_nat <- exp(median_anc_log + 0.5 * diff_unrel_vs_anc) -
  exp(median_anc_log - 0.5 * diff_unrel_vs_anc)


# STEP 6: Assemble posterior draws for plotting


df_draws <- tibble(
  contrast = factor(
    rep(
      c(
        "Reliable - Ancestral",
        "Unreliable - Ancestral"
      ),
      each = length(diff_rel_nat)
    ),
    levels = c(
        "Reliable - Ancestral",
        "Unreliable - Ancestral"
    )
  ),
  diff_fit = c(diff_rel_nat, diff_unrel_nat)
)


# STEP 7: Summarize posterior distributions


df_summ <- df_draws %>%
  group_by(contrast) %>%
  summarise(
    med = median(diff_fit),
    lo = quantile(diff_fit, growth_lo_p),
    hi = quantile(diff_fit, growth_hi_p),
    .groups = "drop"
  )


# STEP 8: Plot (facet label at top, consistent style)


growth_plot_halfeye_with_summary(
  df_draws = df_draws %>% mutate(x = " "),
  df_summ  = df_summ  %>% mutate(x = " "),
  x = "x",
  y = "diff_fit",
  x_label = NULL,
  y_label = "Adaptive Divergence",
  hline_value = 0,
  single_fill = unname(contrast_color)
) +
  facet_wrap(~ contrast, nrow = 1) +
  boxed_strip_theme()+
  xlab("Experimental Evolution Contrast")
```


Showing the average log fitness transformed to the natural R0 scale. i.e. geo mean of R0



```{r growth_plot_average_growth_custom_blue_rel_all_unrel, fig.width=8, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}

# Build treatment-specific log-fitness values

alpha_anc <- growth_draws_df$alpha_anc
alpha_rel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[1]`
alpha_unrel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[2]`

state_df_anc <- growth_draws_df %>%
  select(paste0("state_anc[", 1:8, "]")) %>%
  setNames(1:8)

state_df_rel_raw <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[1,", 1:8, "]")
  )

state_df_unrel_raw <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[2,", 1:8, "]")
  )

log_fit_anc <- (state_df_anc + alpha_anc) %>%
  as_tibble() %>%
  setNames(1:8)

log_fit_rel <- (state_df_rel_raw[1:8] + state_df_rel_raw[9:16] + alpha_rel) %>%
  as_tibble() %>%
  setNames(1:8)

log_fit_unrel <- (state_df_unrel_raw[1:8] + state_df_unrel_raw[9:16] + alpha_unrel) %>%
  as_tibble() %>%
  setNames(1:8)

# Geometric mean fitness

blue_state_ids <- c(2, 4, 6, 8)
all_state_ids  <- 1:8

avg_anc_blue   <- exp(rowMeans(log_fit_anc[, blue_state_ids]))
avg_rel_blue   <- exp(rowMeans(log_fit_rel[, blue_state_ids]))

avg_anc_all    <- exp(rowMeans(log_fit_anc[, all_state_ids]))
avg_unrel_all  <- exp(rowMeans(log_fit_unrel[, all_state_ids]))

# Posterior draws

n_draw <- nrow(growth_draws_df)

df_draws <- tibble(
  group = factor(
    rep(
      c(
        "Anc (blue only)",
        "Rel (blue only)",
        "Anc (all env)",
        "Unrel (all env)"
      ),
      each = n_draw
    ),
    levels = c(
      "Anc (blue only)",
      "Rel (blue only)",
      "Anc (all env)",
      "Unrel (all env)"
    )
  ),
  x_dummy = factor("", levels = ""),
  mean_growth = c(avg_anc_blue, avg_rel_blue, avg_anc_all, avg_unrel_all)
)

# Summaries

df_summ <- df_draws %>%
  group_by(group, x_dummy) %>%
  summarise(
    med = median(mean_growth),
    lo = quantile(mean_growth, growth_lo_p),
    hi = quantile(mean_growth, growth_hi_p),
    .groups = "drop"
  )

# Plot

growth_plot_halfeye_with_summary(
  df_draws = df_draws,
  df_summ = df_summ,
  x = "x_dummy",
  y = "mean_growth",
  x_label = NULL,
  y_label = "Geometric Mean Fitness",
  hline_value = NULL,
  remove_x_text = TRUE,
  single_fill = unname(contrast_color)
) +
  facet_wrap(
    ~ group,
    nrow = 1,
    labeller = as_labeller(c(
      "Anc (blue only)"  = "Ancestral",
      "Rel (blue only)"  = "Reliable",
      "Anc (all env)"    = "Ancestral",
      "Unrel (all env)"  = "Unreliable"
    ))
  ) +
  geom_text(
    data = data.frame(
      group = c("Anc (blue only)", "Rel (blue only)",
                "Anc (all env)", "Unrel (all env)"),
      x_dummy = "",
      y = -0.1,
      label = c("Blue Light Only", "Blue Light Only",
                "All Environments", "All Environments")
    ),
    aes(x = x_dummy, y = y, label = label),
    inherit.aes = FALSE,
    size = 3
  ) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  boxed_strip_theme(12) +
  theme(
    legend.position = "none",
    axis.line.x = element_blank(),
    plot.margin = margin(t = 5.5, r = 5.5, b = 20, l = 5.5)
  ) +
  scale_y_continuous(breaks = c(0, 1, 2, 3)) +
  coord_cartesian(ylim = c(-0.25, 3))

```


 
Here we go to the natural scale and then take the contrast between cue and no cue R0 averaged across mat ox states


 
```{r growth_plot_overall_cue_contrast_averaged_across_mat, fig.width=9, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}

alpha_anc <- growth_draws_df$alpha_anc
alpha_rel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[1]`
alpha_unrel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[2]`

state_df_anc <- growth_draws_df %>%
  select(paste0("state_anc[", 1:8, "]")) %>%
  setNames(1:8)

state_df_rel_raw <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[1,", 1:8, "]")
  )

state_df_unrel_raw <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[2,", 1:8, "]")
  )

log_fit_anc <- (state_df_anc + alpha_anc) %>%
  as_tibble() %>%
  setNames(1:8)

log_fit_rel <- (state_df_rel_raw[1:8] + state_df_rel_raw[9:16] + alpha_rel) %>%
  as_tibble() %>%
  setNames(1:8)

log_fit_unrel <- (state_df_unrel_raw[1:8] + state_df_unrel_raw[9:16] + alpha_unrel) %>%
  as_tibble() %>%
  setNames(1:8)


 


fit_anc <- exp(log_fit_anc)
fit_rel <- exp(log_fit_rel)
fit_unrel <- exp(log_fit_unrel)



# State pairs are:
#   2 - 1
#   4 - 3
#   6 - 5
#   8 - 7


cue_effect_anc <- (
  (fit_anc$`2` - fit_anc$`1`) +
  (fit_anc$`4` - fit_anc$`3`) +
  (fit_anc$`6` - fit_anc$`5`) +
  (fit_anc$`8` - fit_anc$`7`)
) / 4

cue_effect_rel <- (
  (fit_rel$`2` - fit_rel$`1`) +
  (fit_rel$`4` - fit_rel$`3`) +
  (fit_rel$`6` - fit_rel$`5`) +
  (fit_rel$`8` - fit_rel$`7`)
) / 4

cue_effect_unrel <- (
  (fit_unrel$`2` - fit_unrel$`1`) +
  (fit_unrel$`4` - fit_unrel$`3`) +
  (fit_unrel$`6` - fit_unrel$`5`) +
  (fit_unrel$`8` - fit_unrel$`7`)
) / 4


 

n_draw <- nrow(growth_draws_df)

df_draws <- tibble(
  treatment_box = factor(
    rep(c("Ancestral", "Reliable", "Unreliable"), each = n_draw),
    levels = c("Ancestral", "Reliable", "Unreliable")
  ),
  x_dummy = factor("", levels = ""),
  diff_fit = c(cue_effect_anc, cue_effect_rel, cue_effect_unrel)
)

 


df_summ <- df_draws %>%
  group_by(treatment_box, x_dummy) %>%
  summarise(
    med = median(diff_fit),
    lo = quantile(diff_fit, growth_lo_p),
    hi = quantile(diff_fit, growth_hi_p),
    .groups = "drop"
  )





growth_plot_halfeye_with_summary(
  df_draws = df_draws,
  df_summ = df_summ,
  x = "x_dummy",
  y = "diff_fit",
  x_label = NULL,
  y_label = "Light-Plasticity for Fitness",
  hline_value = NULL,
  remove_x_text = TRUE,
  single_fill = unname(contrast_color)
) +
  facet_wrap(~ treatment_box, nrow = 1) +
  boxed_strip_theme(12) +
  theme(
    legend.position = "none"
  )+ labs(x = "Experimental Evolution Regime")
  
```



Difference in cue effect on R0 between Rel and Unrel


```{r growth_plot_rel_vs_unrel_overall_cue_contrast, fig.width=6, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}

alpha_rel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[1]`
alpha_unrel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[2]`

state_df_rel_raw <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[1,", 1:8, "]")
  )

state_df_unrel_raw <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[2,", 1:8, "]")
  )

log_fit_rel <- (state_df_rel_raw[1:8] + state_df_rel_raw[9:16] + alpha_rel) %>%
  as_tibble() %>%
  setNames(1:8)

log_fit_unrel <- (state_df_unrel_raw[1:8] + state_df_unrel_raw[9:16] + alpha_unrel) %>%
  as_tibble() %>%
  setNames(1:8)

fit_rel <- exp(log_fit_rel)
fit_unrel <- exp(log_fit_unrel)

cue_effect_rel <- (
  (fit_rel$`2` - fit_rel$`1`) +
    (fit_rel$`4` - fit_rel$`3`) +
    (fit_rel$`6` - fit_rel$`5`) +
    (fit_rel$`8` - fit_rel$`7`)
) / 4

cue_effect_unrel <- (
  (fit_unrel$`2` - fit_unrel$`1`) +
    (fit_unrel$`4` - fit_unrel$`3`) +
    (fit_unrel$`6` - fit_unrel$`5`) +
    (fit_unrel$`8` - fit_unrel$`7`)
) / 4

diff_fit <- cue_effect_rel - cue_effect_unrel

df_draws <- tibble(
  contrast_box = factor(
    "Reliable - Unreliable",
    levels = "Reliable - Unreliable"
  ),
  x_dummy = factor("", levels = ""),
  diff_fit = diff_fit
)

df_summ <- df_draws %>%
  group_by(contrast_box, x_dummy) %>%
  summarise(
    med = median(diff_fit),
    lo = quantile(diff_fit, growth_lo_p),
    hi = quantile(diff_fit, growth_hi_p),
    .groups = "drop"
  )

growth_plot_halfeye_with_summary(
  df_draws = df_draws,
  df_summ = df_summ,
  x = "x_dummy",
  y = "diff_fit",
  x_label = NULL,
  y_label = "Light-Plasticity Fitness Differentiation",
  hline_value = 0,
  remove_x_text = TRUE,
  single_fill = unname(contrast_color)
) +
  facet_wrap(~ contrast_box, nrow = 1) +
  boxed_strip_theme(12) +
  labs(
    x = "Experimental Evolution Regime",
    tag = "Overall \n\n Sup Fig. 7"
  ) +
  theme(
    legend.position = "none",
    plot.tag.position = c(0.98, 0.98),
    plot.tag = element_text(
      size = 12,
      face = "bold",
      hjust = 1,
      vjust = 1
    ),
    plot.margin = margin(t = 8, r = 18, b = 5, l = 5)
  )
```



Figure out exactly how much of the posterior is above 0. 6.3% is below zero.
```{r}
df_draws %>%
  group_by(contrast_box, x_dummy) %>%
  summarise(
    med = median(diff_fit),
    lo = quantile(diff_fit, 0.063),
    .groups = "drop"
  )
```



Analyze the total difference in fitness under the blue light cue between Reliable and Unreliable:
```{r growth_plot_rel_vs_unrel_blue_light_diff , echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12 }
alpha_rel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[1]`
alpha_unrel <- growth_draws_df$alpha_anc + growth_draws_df$`delta_alpha[2]`

# Build natural-scale Reliable fitness for each of the 8 state combinations.
state_df_rel <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[1,", 1:8, "]")
  )

state_df_rel_sum <- (state_df_rel[1:8] + state_df_rel[9:16] + alpha_rel) %>%
  as_tibble() %>%
  mutate(across(everything(), exp))

names(state_df_rel_sum) <- as.character(1:8)

# Build natural-scale Unreliable fitness for each of the 8 state combinations.
state_df_unrel <- growth_draws_df %>%
  select(
    paste0("state_anc[", 1:8, "]"),
    paste0("delta_state[2,", 1:8, "]")
  )

state_df_unrel_sum <- (state_df_unrel[1:8] + state_df_unrel[9:16] + alpha_unrel) %>%
  as_tibble() %>%
  mutate(across(everything(), exp))

names(state_df_unrel_sum) <- as.character(1:8)

# Convert Reliable predictions to long format and attach maternal-environment labels.
draws_rel <- state_df_rel_sum %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_rel"
  ) %>%
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id")

# Do the same for Unreliable predictions.
draws_unrel <- state_df_unrel_sum %>%
  mutate(.draw = row_number()) %>%
  pivot_longer(
    cols = all_of(as.character(1:8)),
    names_to = "state",
    values_to = "R0_unrel"
  ) %>%
  mutate(
    state_id = as.integer(state)
  ) %>%
  left_join(state_key_growth, by = "state_id")

# Keep only Blue Light states, then contrast Reliable and Unreliable fitness
# within each posterior draw and maternal environment.
df_draws <- draws_rel %>%
  filter(cue == "Blue Light") %>%
  select(.draw, mat_label, R0_rel) %>%
  left_join(
    draws_unrel %>%
      filter(cue == "Blue Light") %>%
      select(.draw, mat_label, R0_unrel),
    by = c(".draw", "mat_label")
  ) %>%
  mutate(
    rel_minus_unrel_blue_fitness = R0_rel - R0_unrel,
    mat_label = factor(mat_label, levels = growth_key_mat_levels)
  )

# Summarize the posterior distribution for each maternal environment.
df_summ <- df_draws %>%
  group_by(mat_label) %>%
  summarise(
    med = median(rel_minus_unrel_blue_fitness),
    lo = quantile(rel_minus_unrel_blue_fitness, growth_lo_p),
    hi = quantile(rel_minus_unrel_blue_fitness, growth_hi_p),
    .groups = "drop"
  )

# Plot posterior half-eyes, medians, and intervals.
growth_plot_halfeye_with_summary(
  df_draws = df_draws,
  df_summ = df_summ,
  x = "mat_label",
  y = "rel_minus_unrel_blue_fitness",
  x_label = axis_label_mat_env_full,
  y_label = "Blue Light Fitness Divergence",
  hline_value = 0,
  show_legend = FALSE,
  single_fill = unname(contrast_color)
)
```







```{r growth_plot_growth_variance_pairs_shared_pop, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=8}

xlim_common <- c(0, 1.2)
ylim_common <- c(0, 1.2)

draws_var <- growth_draws_df %>%
  transmute(
    `Block SD` = sigma_block,
    `Sequence SD` = sigma_seq,
    `Replicate Population SD` = sigma_pop,
    `Likelihood SD` = sigma_y
  )

draw_variance_matrix <- function(df_plot) {
  vars <- names(df_plot)
  k <- length(vars)

  grid::grid.newpage()
  grid::pushViewport(grid::viewport(layout = grid::grid.layout(k, k)))

  for (i in seq_len(k)) {
    for (j in seq_len(k)) {

      grid::pushViewport(
        grid::viewport(layout.pos.row = i, layout.pos.col = j)
      )

      xvar <- vars[j]
      yvar <- vars[i]

      if (i == j) {
        p <- ggplot(df_plot, aes(x = .data[[xvar]])) +
          geom_density(fill = "steelblue", alpha = 0.7, color = "navy") +
          coord_cartesian(xlim = xlim_common) +
          labs(
            x = xvar,
            y = "Density"
          ) +scale_x_continuous(n.breaks = 3)+
scale_y_continuous(n.breaks = 3)+
          theme_bw(base_size = 14) +
          theme(
            axis.title.x = element_text(size = 12),
  axis.title.y = element_text(size = 12),
  axis.text.x  = element_text(size = 8),
  axis.text.y  = element_text(size = 8)
          )
      } else {
        p <- ggplot(df_plot, aes(x = .data[[xvar]], y = .data[[yvar]])) +
          geom_point(size = 0.5, alpha = 0.5, color = "navy") +
          coord_cartesian(xlim = xlim_common, ylim = ylim_common) +
          labs(
            x = xvar,
            y = yvar
          ) +scale_x_continuous(n.breaks = 3)+
scale_y_continuous(n.breaks = 3)+
          theme_bw(base_size = 12) +
          theme(
            axis.title.x = element_text(size = 12),
  axis.title.y = element_text(size = 12),
  axis.text.x  = element_text(size = 8),
  axis.text.y  = element_text(size = 8)
          )
      }

      print(p, newpage = FALSE)
      grid::popViewport()
    }
  }

  grid::popViewport()
}

draw_variance_matrix(draws_var)
```





### Sequence specific fitness estimates






And now for the model with both Rel and Unrel fit at the same time.

```{r growth_read_draws_anc_rel_unrel_by_seq_state_only, include=FALSE}
draws_growth_anc_rel_unrel_by_seq_state_only <- readRDS(
  "stan_output/growth_anc_rel_unrel_by_seq_state_only_draws.rds"
)

growth_anc_rel_unrel_by_seq_draws_df <- posterior::as_draws_df(
  draws_growth_anc_rel_unrel_by_seq_state_only
)
```


 
```{r growth_build_plot_df_rel_by_sequence_state_only_from_full_model, include=FALSE}
# Extract posterior draws of sequence-specific Reliable mean log fitness
# for each state, then transform to the natural scale for plotting.

plot_df_rel_seq_from_full_model <- growth_anc_rel_unrel_by_seq_draws_df %>%
  dplyr::select(dplyr::starts_with("mu_rel_state[")) %>%
  tibble::rowid_to_column(".draw") %>%
  tidyr::pivot_longer(
    cols = -.draw,
    names_to = "param",
    values_to = "log_growth"
  ) %>%
  tidyr::extract(
    param,
    into = c("seq_id", "state_id"),
    regex = "mu_rel_state\\[(\\d+),(\\d+)\\]",
    convert = TRUE
  ) %>%
  dplyr::mutate(
    growth = exp(log_growth)
  ) %>%
  dplyr::left_join(state_key_growth, by = "state_id") %>%
  dplyr::mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light")),
    sequence = factor(
      paste("Sequence", seq_id),
      levels = paste("Sequence", sort(unique(seq_id)))
    )
  )

plot_summ_rel_seq_from_full_model <- plot_df_rel_seq_from_full_model %>%
  dplyr::group_by(sequence, mat_label, cue) %>%
  dplyr::summarise(
    med = median(growth),
    lo = quantile(growth, growth_lo_p),
    hi = quantile(growth, growth_hi_p),
    .groups = "drop"
  )
```


 
```{r growth_plot_rel_by_sequence_matcue_intervals_state_only_from_full_model, fig.width=9, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  plot_summ_rel_seq_from_full_model,
  aes(x = mat_label, y = med, fill = cue)
) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.8
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.9),
    width = 0.2,
    linewidth = 0.7
  ) +
  facet_wrap(~ sequence, nrow = 1) +
  scale_fill_manual(
    name = legend_label_light_cue,
    values = growth_cue_plot_colors
  ) +
  labs(
    x = axis_label_mat_env_full,
    y = axis_label_growth
  ) +
  growth_base_plot_theme()
```




```{r growth_build_plot_df_unrel_by_sequence_state_only_from_full_model, include=FALSE}
# Extract posterior draws of sequence-specific Unreliable mean log fitness
# for each state, then transform to the natural scale for plotting.

plot_df_unrel_seq_from_full_model <- growth_anc_rel_unrel_by_seq_draws_df %>%
  dplyr::select(dplyr::starts_with("mu_unrel_state[")) %>%
  tibble::rowid_to_column(".draw") %>%
  tidyr::pivot_longer(
    cols = -.draw,
    names_to = "param",
    values_to = "log_growth"
  ) %>%
  tidyr::extract(
    param,
    into = c("seq_id", "state_id"),
    regex = "mu_unrel_state\\[(\\d+),(\\d+)\\]",
    convert = TRUE
  ) %>%
  dplyr::mutate(
    growth = exp(log_growth)
  ) %>%
  dplyr::left_join(state_key_growth, by = "state_id") %>%
  dplyr::mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light")),
    sequence = factor(
      paste("Sequence", seq_id),
      levels = paste("Sequence", sort(unique(seq_id)))
    )
  )

plot_summ_unrel_seq_from_full_model <- plot_df_unrel_seq_from_full_model %>%
  dplyr::group_by(sequence, mat_label, cue) %>%
  dplyr::summarise(
    med = median(growth),
    lo = quantile(growth, growth_lo_p),
    hi = quantile(growth, growth_hi_p),
    .groups = "drop"
  )
```



```{r growth_plot_unrel_by_sequence_matcue_intervals_state_only_from_full_model, fig.width=9, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  plot_summ_unrel_seq_from_full_model,
  aes(x = mat_label, y = med, fill = cue)
) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.8
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.9),
    width = 0.2,
    linewidth = 0.7
  ) +
  facet_wrap(~ sequence, nrow = 1) +
  scale_fill_manual(
    name = legend_label_light_cue,
    values = growth_cue_plot_colors
  ) +
  labs(
    x = axis_label_mat_env_full,
    y = axis_label_growth
  ) +
  growth_base_plot_theme()
```



Plot the fitness divergence between Reliable and Unreliable by sequence. To do this we average over maternal oxygen envirnoments and only take the contrast the presence of blue light.
```{r growth_build_plot_df_rel_minus_unrel_blue_only_by_sequence_from_full_model, include=FALSE}


plot_df_rel_blue_only_by_seq <- growth_anc_rel_unrel_by_seq_draws_df %>%
  dplyr::select(dplyr::starts_with("mu_rel_state[")) %>%
  tibble::rowid_to_column(".draw") %>%
  tidyr::pivot_longer(
    cols = -.draw,
    names_to = "param",
    values_to = "log_growth"
  ) %>%
  tidyr::extract(
    param,
    into = c("seq_id", "state_id"),
    regex = "mu_rel_state\\[(\\d+),(\\d+)\\]",
    convert = TRUE
  ) %>%
  dplyr::mutate(
    rel_growth = exp(log_growth)
  ) %>%
  dplyr::select(.draw, seq_id, state_id, rel_growth) %>%
  dplyr::left_join(state_key_growth, by = "state_id") %>%
  dplyr::filter(cue == "Blue Light")

plot_df_unrel_blue_only_by_seq <- growth_anc_rel_unrel_by_seq_draws_df %>%
  dplyr::select(dplyr::starts_with("mu_unrel_state[")) %>%
  tibble::rowid_to_column(".draw") %>%
  tidyr::pivot_longer(
    cols = -.draw,
    names_to = "param",
    values_to = "log_growth"
  ) %>%
  tidyr::extract(
    param,
    into = c("seq_id", "state_id"),
    regex = "mu_unrel_state\\[(\\d+),(\\d+)\\]",
    convert = TRUE
  ) %>%
  dplyr::mutate(
    unrel_growth = exp(log_growth)
  ) %>%
  dplyr::select(.draw, seq_id, state_id, unrel_growth) %>%
  dplyr::left_join(state_key_growth, by = "state_id") %>%
  dplyr::filter(cue == "Blue Light")

plot_df_rel_minus_unrel_blue_only_by_seq <- plot_df_rel_blue_only_by_seq %>%
  dplyr::select(.draw, seq_id, state_id, rel_growth) %>%
  dplyr::left_join(
    plot_df_unrel_blue_only_by_seq %>%
      dplyr::select(.draw, seq_id, state_id, unrel_growth),
    by = c(".draw", "seq_id", "state_id")
  ) %>%
  dplyr::group_by(.draw, seq_id) %>%
  dplyr::summarise(
    diff_rel_minus_unrel = mean(rel_growth - unrel_growth),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    sequence = factor(
      paste("Sequence", seq_id),
      levels = paste("Sequence", sort(unique(seq_id)))
    )
  )

plot_summ_rel_minus_unrel_blue_only_by_seq <- plot_df_rel_minus_unrel_blue_only_by_seq %>%
  dplyr::group_by(sequence) %>%
  dplyr::summarise(
    med = median(diff_rel_minus_unrel),
    lo = quantile(diff_rel_minus_unrel, growth_lo_p),
    hi = quantile(diff_rel_minus_unrel, growth_hi_p),
    .groups = "drop"
  )
```

```{r growth_plot_rel_minus_unrel_blue_only_by_sequence_from_full_model, echo=FALSE, message=FALSE, warning=FALSE}
growth_plot_halfeye_with_summary(
  df_draws = plot_df_rel_minus_unrel_blue_only_by_seq,
  df_summ = plot_summ_rel_minus_unrel_blue_only_by_seq,
  x = "sequence",
  y = "diff_rel_minus_unrel",
  x_label = NULL,
  y_label = "Blue Light Fitness Divergence",
  single_fill = contrast_color
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.4,
    color = "grey40"
  ) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

Get prepared to plot the cue response divergence by maternal env for each seq.
```{r growth_build_plot_df_rel_minus_unrel_log_cue_contrast_by_sequence_from_full_model, include=FALSE}
# Build posterior draw-level contrasts of the Reliable cue response
# against the Unreliable cue response, separately for each sequence
# and maternal environment, on the log scale.
#
# For each draw, sequence, and maternal environment:
#   cue effect = log_growth_BlueLight - log_growth_NoLight
#   contrast   = Reliable cue effect - Unreliable cue effect

plot_df_rel_seq_log_cue_only <- growth_anc_rel_unrel_by_seq_draws_df %>%
  dplyr::select(dplyr::starts_with("mu_rel_state[")) %>%
  tibble::rowid_to_column(".draw") %>%
  tidyr::pivot_longer(
    cols = -.draw,
    names_to = "param",
    values_to = "log_growth"
  ) %>%
  tidyr::extract(
    param,
    into = c("seq_id", "state_id"),
    regex = "mu_rel_state\\[(\\d+),(\\d+)\\]",
    convert = TRUE
  ) %>%
  dplyr::left_join(state_key_growth, by = "state_id") %>%
  dplyr::mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  ) %>%
  dplyr::select(.draw, seq_id, mat_label, cue, log_growth) %>%
  tidyr::pivot_wider(
    names_from = cue,
    values_from = log_growth
  ) %>%
  dplyr::mutate(
    cue_contrast_rel_log = exp(`Blue Light`) - exp(`No Light`)
  ) %>%
  dplyr::select(.draw, seq_id, mat_label, cue_contrast_rel_log)

plot_df_unrel_seq_log_cue_only <- growth_anc_rel_unrel_by_seq_draws_df %>%
  dplyr::select(dplyr::starts_with("mu_unrel_state[")) %>%
  tibble::rowid_to_column(".draw") %>%
  tidyr::pivot_longer(
    cols = -.draw,
    names_to = "param",
    values_to = "log_growth"
  ) %>%
  tidyr::extract(
    param,
    into = c("seq_id", "state_id"),
    regex = "mu_unrel_state\\[(\\d+),(\\d+)\\]",
    convert = TRUE
  ) %>%
  dplyr::left_join(state_key_growth, by = "state_id") %>%
  dplyr::mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  ) %>%
  dplyr::select(.draw, seq_id, mat_label, cue, log_growth) %>%
  tidyr::pivot_wider(
    names_from = cue,
    values_from = log_growth
  ) %>%
  dplyr::mutate(
    cue_contrast_unrel_log = exp(`Blue Light`) - exp(`No Light`)
  ) %>%
  dplyr::select(.draw, seq_id, mat_label, cue_contrast_unrel_log)

plot_df_rel_minus_unrel_log_cue_contrast_by_seq <- plot_df_rel_seq_log_cue_only %>%
  dplyr::left_join(
    plot_df_unrel_seq_log_cue_only,
    by = c(".draw", "seq_id", "mat_label")
  ) %>%
  dplyr::mutate(
    sequence = factor(
      paste("Sequence", seq_id),
      levels = paste("Sequence", sort(unique(seq_id)))
    ),
    rel_minus_unrel_log_cue_contrast =
      cue_contrast_rel_log - cue_contrast_unrel_log
  )

plot_summ_rel_minus_unrel_log_cue_contrast_by_seq <- plot_df_rel_minus_unrel_log_cue_contrast_by_seq %>%
  dplyr::group_by(sequence, mat_label) %>%
  dplyr::summarise(
    med = median(rel_minus_unrel_log_cue_contrast),
    lo = quantile(rel_minus_unrel_log_cue_contrast, growth_lo_p),
    hi = quantile(rel_minus_unrel_log_cue_contrast, growth_hi_p),
    .groups = "drop"
  )
```


This is the contrast in cue response between reliable and unreliable on the log scale. 


```{r growth_plot_rel_minus_unrel_log_cue_contrast_by_sequence_from_full_model, fig.width=7, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}
growth_plot_halfeye_with_summary(
  df_draws = plot_df_rel_minus_unrel_log_cue_contrast_by_seq,
  df_summ = plot_summ_rel_minus_unrel_log_cue_contrast_by_seq,
  x = "mat_label",
  y = "rel_minus_unrel_log_cue_contrast",
  x_label = axis_label_mat_env_full,
  y_label = "Cue Response Divergence",
  hline_value = 0,
  single_fill = unname(contrast_color)
) +
  facet_wrap(~ sequence, nrow = 1) +
  boxed_strip_theme(12)+
  ylim(-2,2)
```

```{r growth_build_plot_df_rel_minus_unrel_log_cue_contrast_average_from_full_model, include=FALSE}
# Average the Reliable - Unreliable cue-response contrast
# across both sequence and maternal environment within each draw.

plot_df_rel_minus_unrel_log_cue_contrast_avg <- plot_df_rel_minus_unrel_log_cue_contrast_by_seq %>%
  dplyr::group_by(.draw) %>%
  dplyr::summarise(
    rel_minus_unrel_log_cue_contrast_avg =
      mean(rel_minus_unrel_log_cue_contrast),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    contrast_label = factor(
      "Reliable - Unreliable",
      levels = "Reliable - Unreliable"
    )
  )

plot_summ_rel_minus_unrel_log_cue_contrast_avg <- plot_df_rel_minus_unrel_log_cue_contrast_avg %>%
  dplyr::group_by(contrast_label) %>%
  dplyr::summarise(
    med = median(rel_minus_unrel_log_cue_contrast_avg),
    lo = quantile(rel_minus_unrel_log_cue_contrast_avg, growth_lo_p),
    hi = quantile(rel_minus_unrel_log_cue_contrast_avg, growth_hi_p),
    .groups = "drop"
  )
```



```{r growth_plot_rel_minus_unrel_log_cue_contrast_average_from_full_model, fig.width=4.8, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}
growth_plot_halfeye_with_summary(
  df_draws = plot_df_rel_minus_unrel_log_cue_contrast_avg,
  df_summ = plot_summ_rel_minus_unrel_log_cue_contrast_avg,
  x = "contrast_label",
  y = "rel_minus_unrel_log_cue_contrast_avg",
  x_label = NULL,
  y_label = "Average Reliable cue response - Unreliable cue response",
  hline_value = 0,
  single_fill = unname(contrast_color)
)
```

now figure out exactly how much is below 0. The answer is 7.9%.
```{r}
plot_df_rel_minus_unrel_log_cue_contrast_avg %>%
  dplyr::group_by(contrast_label) %>%
  dplyr::summarise(
    med = median(rel_minus_unrel_log_cue_contrast_avg),
    lo = quantile(rel_minus_unrel_log_cue_contrast_avg, 0.079),
    hi = quantile(rel_minus_unrel_log_cue_contrast_avg, growth_hi_p),
    .groups = "drop"
  )
```


## Fecundity data



```{r fec_plot_anc_fec_by_matcue_intervals_from_mat_and_cue_only, fig.width=6, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}

alpha_fix <- median(fec_draws_df$`alpha[1]`)

mat_df <- fec_draws_df %>%
  select(starts_with("mat_eff[1,")) %>%
  setNames(1:4) %>%
  as.matrix()

beta_cue_anc <- fec_draws_df$`beta_cue[1]`

plot_df <- tidyr::expand_grid(
  .draw = seq_len(nrow(mat_df)),
  mat_id = 1:4,
  cue = c("No Light", "Blue Light")
) %>%
  mutate(
    Ci = if_else(cue == "Blue Light", 1, -1),
    eta = alpha_fix +
      mat_df[cbind(.draw, mat_id)] +
      beta_cue_anc[.draw] * Ci,
    fec = exp(eta)
  ) %>%
  left_join(mat_key_fec, by = "mat_id") %>%
  mutate(
    mat = factor(mat, levels = c("AN", "NN", "AA", "NA"))
  ) %>%
  group_by(mat, cue) %>%
  summarise(
    med = median(fec),
    lo = quantile(fec, fec_lo_p),
    hi = quantile(fec, fec_hi_p),
    .groups = "drop"
  )

ggplot(plot_df, aes(x = mat, y = med, fill = cue)) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.8
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.9),
    width = 0.2,
    linewidth = 0.7
  ) +
  scale_fill_manual(
    name = legend_label_light_cue,
    values = fec_cue_plot_colors
  ) +
  labs(
    x = axis_label_mat_env_full,
    y = axis_label_fecundity
  ) +
  growth_base_plot_theme()
```


```{r plot_anc_growth_and_fec_by_matcue_two_panel, fig.width=10, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}
library(patchwork)

# Standardize key column names for plotting
state_key_growth_plot <- state_key_growth
if ("mat_label" %in% names(state_key_growth_plot) && !("mat" %in% names(state_key_growth_plot))) {
  state_key_growth_plot <- state_key_growth_plot %>%
    rename(mat = mat_label)
}

mat_key_fec_plot <- mat_key_fec
if ("mat_label" %in% names(mat_key_fec_plot) && !("mat" %in% names(mat_key_fec_plot))) {
  mat_key_fec_plot <- mat_key_fec_plot %>%
    rename(mat = mat_label)
}

 
# Left panel: ancestral growth 
alpha_fix_growth <- median(growth_draws_df$alpha_anc)

state_df_growth <- growth_draws_df %>%
  select(starts_with("state_anc[")) %>%
  mutate(across(everything(), ~ exp(alpha_fix_growth + .)))

df_summ_growth <- state_df_growth %>%
  pivot_longer(
    cols = everything(),
    names_to = "state",
    values_to = "grow"
  ) %>%
  mutate(
    state_id = readr::parse_number(state)
  ) %>%
  left_join(state_key_growth_plot, by = "state_id") %>%
  mutate(
    mat = factor(mat, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  ) %>%
  group_by(mat, cue) %>%
  summarise(
    med = median(grow),
    lo = quantile(grow, growth_lo_p),
    hi = quantile(grow, growth_hi_p),
    .groups = "drop"
  )

p_growth <- ggplot(df_summ_growth, aes(x = mat, y = med, fill = cue)) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.8
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.9),
    width = 0.2,
    linewidth = 0.7
  ) +
  scale_fill_manual(
    name = legend_label_light_cue,
    values = growth_cue_plot_colors
  ) +
  labs(
    title = NULL,
    x = axis_label_mat_env_full,
    y = axis_label_growth
  ) +
  growth_base_plot_theme() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
 
# Right panel: ancestral fecundity

alpha_fix_fec <- median(fec_draws_df$`alpha[1]`)

mat_df_fec <- fec_draws_df %>%
  select(starts_with("mat_eff[1,")) %>%
  setNames(1:4) %>%
  as.matrix()

beta_cue_anc <- fec_draws_df$`beta_cue[1]`

df_summ_fec <- tidyr::expand_grid(
  .draw = seq_len(nrow(mat_df_fec)),
  mat_id = 1:4,
  cue = c("No Light", "Blue Light")
) %>%
  mutate(
    Ci = if_else(cue == "Blue Light", 1, -1),
    eta = alpha_fix_fec +
      mat_df_fec[cbind(.draw, mat_id)] +
      beta_cue_anc[.draw] * Ci,
    fec = exp(eta)
  ) %>%
  left_join(mat_key_fec_plot, by = "mat_id") %>%
  mutate(
    mat = factor(mat, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  ) %>%
  group_by(mat, cue) %>%
  summarise(
    med = median(fec),
    lo = quantile(fec, fec_lo_p),
    hi = quantile(fec, fec_hi_p),
    .groups = "drop"
  )

p_fec <- ggplot(df_summ_fec, aes(x = mat, y = med, fill = cue)) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.8
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.9),
    width = 0.2,
    linewidth = 0.7
  ) +
  scale_fill_manual(
    name = legend_label_light_cue,
    values = fec_cue_plot_colors
  ) +
  labs(NULL,
    x = axis_label_mat_env_full,
    y = axis_label_fecundity
  ) +
  fec_base_plot_theme() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )


# Combine panels


p_growth + p_fec + plot_layout(ncol = 2, guides = "collect") &
  theme(legend.position = "right")
```






```{r fec_plot_all_fec_by_matcue_intervals_from_mat_and_cue_only, fig.width=9, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}
# Predicted fecundity for each maternal environment x cue combination
# for all three treatments from the additive fecundity model.
#
# The treatment-specific linear predictor is:
#   eta = alpha[t] + mat_eff[t, mat_id] + beta_cue[t] * Ci
#
# We fix each treatment intercept at its posterior median, then combine it
# with posterior draws of the treatment-specific maternal and cue effects,
# and transform back to the natural scale with exp().
 

treatment_key_fec <- tibble::tibble(
  treatment_id = 1:3,
  treatment = factor(
    c("Ancestral", "Reliable", "Unreliable"),
    levels = c("Ancestral", "Reliable", "Unreliable")
  )
)

mat_key_fec_plot <- mat_key_fec
if ("mat_label" %in% names(mat_key_fec_plot) && !("mat" %in% names(mat_key_fec_plot))) {
  mat_key_fec_plot <- mat_key_fec_plot %>%
    rename(mat = mat_label)
}

alpha_fix <- c(
  median(fec_draws_df$`alpha[1]`),
  median(fec_draws_df$`alpha[2]`),
  median(fec_draws_df$`alpha[3]`)
)

mat_df_1 <- fec_draws_df %>%
  select(starts_with("mat_eff[1,")) %>%
  setNames(1:4) %>%
  as.matrix()

mat_df_2 <- fec_draws_df %>%
  select(starts_with("mat_eff[2,")) %>%
  setNames(1:4) %>%
  as.matrix()

mat_df_3 <- fec_draws_df %>%
  select(starts_with("mat_eff[3,")) %>%
  setNames(1:4) %>%
  as.matrix()

beta_cue_1 <- fec_draws_df$`beta_cue[1]`
beta_cue_2 <- fec_draws_df$`beta_cue[2]`
beta_cue_3 <- fec_draws_df$`beta_cue[3]`

plot_df <- tidyr::expand_grid(
  .draw = seq_len(nrow(fec_draws_df)),
  treatment_id = 1:3,
  mat_id = 1:4,
  cue = c("No Light", "Blue Light")
) %>%
  mutate(
    Ci = if_else(cue == "Blue Light", 1, -1),
    mat_val = case_when(
      treatment_id == 1 ~ mat_df_1[cbind(.draw, mat_id)],
      treatment_id == 2 ~ mat_df_2[cbind(.draw, mat_id)],
      treatment_id == 3 ~ mat_df_3[cbind(.draw, mat_id)]
    ),
    beta_val = case_when(
      treatment_id == 1 ~ beta_cue_1[.draw],
      treatment_id == 2 ~ beta_cue_2[.draw],
      treatment_id == 3 ~ beta_cue_3[.draw]
    ),
    eta = alpha_fix[treatment_id] + mat_val + beta_val * Ci,
    fec = exp(eta)
  ) %>%
  left_join(mat_key_fec_plot, by = "mat_id") %>%
  left_join(treatment_key_fec, by = "treatment_id") %>%
  mutate(
    mat = factor(mat, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  )

plot_summ <- plot_df %>%
  group_by(treatment, mat, cue) %>%
  summarise(
    med = median(fec),
    lo = quantile(fec, fec_lo_p),
    hi = quantile(fec, fec_hi_p),
    .groups = "drop"
  )

ggplot(plot_summ, aes(x = mat, y = med, fill = cue)) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.8
  ) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    position = position_dodge(width = 0.9),
    width = 0.2,
    linewidth = 0.7
  ) +
  facet_wrap(~ treatment, nrow = 1) +
  scale_fill_manual(
    name = legend_label_light_cue,
    values = fec_cue_plot_colors
  ) +
  labs(
    x = axis_label_mat_env_full,
    y = axis_label_fecundity
  ) +
  fec_base_plot_theme()
```






```{r fec_plot_difference_anc_vs_rel_unrel_matcue_only, echo=FALSE, message=FALSE, warning=FALSE}
alpha_anc <- fec_draws_df$`alpha[1]` # ancestral intercept
alpha_rel <- fec_draws_df$`alpha[2]` # reliable intercept
alpha_unrel <- fec_draws_df$`alpha[3]` # unreliable intercept

beta_anc <- fec_draws_df$`beta_cue[1]`
beta_rel <- fec_draws_df$`beta_cue[2]`
beta_unrel <- fec_draws_df$`beta_cue[3]`

mat_key_fec_plot <- mat_key_fec
if ("mat" %in% names(mat_key_fec_plot) && !("mat_label" %in% names(mat_key_fec_plot))) {
  mat_key_fec_plot <- mat_key_fec_plot %>%
    rename(mat_label = mat)
}

# calc anc fecundity from posterior intercept + maternal effect + cue effect
mat_df_anc <- fec_draws_df %>%
  select(starts_with("mat_eff[1,")) %>%
  setNames(as.character(1:4))

draws_anc <- tidyr::expand_grid(
  .draw = seq_len(nrow(fec_draws_df)),
  mat_id = 1:4,
  Ci = c(-1L, 1L)
) %>%
  mutate(
    cue = if_else(Ci == 1L, "Blue Light", "No Light"),
    mat_1 = mat_df_anc[[1]][.draw],
    mat_2 = mat_df_anc[[2]][.draw],
    mat_3 = mat_df_anc[[3]][.draw],
    mat_4 = mat_df_anc[[4]][.draw],
    eta_anc = case_when(
      mat_id == 1L ~ alpha_anc[.draw] + mat_1 + beta_anc[.draw] * Ci,
      mat_id == 2L ~ alpha_anc[.draw] + mat_2 + beta_anc[.draw] * Ci,
      mat_id == 3L ~ alpha_anc[.draw] + mat_3 + beta_anc[.draw] * Ci,
      mat_id == 4L ~ alpha_anc[.draw] + mat_4 + beta_anc[.draw] * Ci
    ),
    fec_anc = exp(eta_anc)
  ) %>%
  left_join(mat_key_fec_plot, by = "mat_id") %>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  ) %>%
  select(.draw, mat_id, Ci, mat_label, cue, fec_anc)

# calc rel fecundity from posterior intercept + maternal effect + cue effect
mat_df_rel <- fec_draws_df %>%
  select(starts_with("mat_eff[2,")) %>%
  setNames(as.character(1:4))

draws_rel <- tidyr::expand_grid(
  .draw = seq_len(nrow(fec_draws_df)),
  mat_id = 1:4,
  Ci = c(-1L, 1L)
) %>%
  mutate(
    cue = if_else(Ci == 1L, "Blue Light", "No Light"),
    mat_1 = mat_df_rel[[1]][.draw],
    mat_2 = mat_df_rel[[2]][.draw],
    mat_3 = mat_df_rel[[3]][.draw],
    mat_4 = mat_df_rel[[4]][.draw],
    eta_rel = case_when(
      mat_id == 1L ~ alpha_rel[.draw] + mat_1 + beta_rel[.draw] * Ci,
      mat_id == 2L ~ alpha_rel[.draw] + mat_2 + beta_rel[.draw] * Ci,
      mat_id == 3L ~ alpha_rel[.draw] + mat_3 + beta_rel[.draw] * Ci,
      mat_id == 4L ~ alpha_rel[.draw] + mat_4 + beta_rel[.draw] * Ci
    ),
    fec_rel = exp(eta_rel)
  ) %>%
  left_join(mat_key_fec_plot, by = "mat_id") %>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  ) %>%
  select(.draw, mat_id, Ci, mat_label, cue, fec_rel)

# calc unrel fecundity from posterior intercept + maternal effect + cue effect
mat_df_unrel <- fec_draws_df %>%
  select(starts_with("mat_eff[3,")) %>%
  setNames(as.character(1:4))

draws_unrel <- tidyr::expand_grid(
  .draw = seq_len(nrow(fec_draws_df)),
  mat_id = 1:4,
  Ci = c(-1L, 1L)
) %>%
  mutate(
    cue = if_else(Ci == 1L, "Blue Light", "No Light"),
    mat_1 = mat_df_unrel[[1]][.draw],
    mat_2 = mat_df_unrel[[2]][.draw],
    mat_3 = mat_df_unrel[[3]][.draw],
    mat_4 = mat_df_unrel[[4]][.draw],
    eta_unrel = case_when(
      mat_id == 1L ~ alpha_unrel[.draw] + mat_1 + beta_unrel[.draw] * Ci,
      mat_id == 2L ~ alpha_unrel[.draw] + mat_2 + beta_unrel[.draw] * Ci,
      mat_id == 3L ~ alpha_unrel[.draw] + mat_3 + beta_unrel[.draw] * Ci,
      mat_id == 4L ~ alpha_unrel[.draw] + mat_4 + beta_unrel[.draw] * Ci
    ),
    fec_unrel = exp(eta_unrel)
  ) %>%
  left_join(mat_key_fec_plot, by = "mat_id") %>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    cue = factor(cue, levels = c("No Light", "Blue Light"))
  ) %>%
  select(.draw, mat_id, Ci, mat_label, cue, fec_unrel)

# put the Anc and Rel into a single df and calc the difference in fecundity
df_draws_rel <- draws_rel %>%
  left_join(
    draws_anc,
    by = c(".draw", "mat_id", "Ci", "mat_label", "cue")
  ) %>%
  mutate(
    diff_fit = fec_rel - fec_anc,
    contrast = factor(
      "Reliable - Ancestral",
      levels = c("Reliable - Ancestral", "Unreliable - Ancestral")
    )
  )

# put the Anc and Unrel into a single df and calc the difference in fecundity
df_draws_unrel <- draws_unrel %>%
  left_join(
    draws_anc,
    by = c(".draw", "mat_id", "Ci", "mat_label", "cue")
  ) %>%
  mutate(
    diff_fit = fec_unrel - fec_anc,
    contrast = factor(
      "Unreliable - Ancestral",
      levels = c("Reliable - Ancestral", "Unreliable - Ancestral")
    )
  )

# combine Rel and Unrel dfs
df_draws_both <- bind_rows(df_draws_rel, df_draws_unrel)

# summarise median and quantiles
df_summ_both <- df_draws_both %>%
  group_by(contrast, mat_label, cue) %>%
  summarise(
    med = median(diff_fit),
    lo = quantile(diff_fit, fec_lo_p),
    hi = quantile(diff_fit, fec_hi_p),
    .groups = "drop"
  )

# display plot
growth_plot_halfeye_with_summary(
  df_draws = df_draws_both,
  df_summ = df_summ_both,
  x = "mat_label",
  y = "diff_fit",
  fill = "cue",
  group = "cue",
  x_label = axis_label_mat_env_full,
  y_label = "Fecundity Divergence",
  show_legend = TRUE,
  dodge_width = 0.7
) +
  facet_wrap(~ contrast, nrow = 1) +
  fill_legend_no_point() +
  boxed_strip_theme(12) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )+
  ylim(-6,6)
```



```{r fec_plot_rel_minus_unrel_mat_only, echo=FALSE, message=FALSE, warning=FALSE}
alpha_fix_rel <- median(fec_draws_df$`alpha[2]`) # fixed reliable intercept
alpha_fix_unrel <- median(fec_draws_df$`alpha[3]`) # fixed unreliable intercept

mat_key_fec_plot <- mat_key_fec
if ("mat" %in% names(mat_key_fec_plot) && !("mat_label" %in% names(mat_key_fec_plot))) {
  mat_key_fec_plot <- mat_key_fec_plot %>%
    rename(mat_label = mat)
}

# reliable maternal effects
mat_df_rel <- fec_draws_df %>%
  select(starts_with("mat_eff[2,")) %>%
  setNames(as.character(1:4)) %>%
  as.matrix()

# unreliable maternal effects
mat_df_unrel <- fec_draws_df %>%
  select(starts_with("mat_eff[3,")) %>%
  setNames(as.character(1:4)) %>%
  as.matrix()

# Maternal-environment-specific Reliable - Unreliable fecundity differences
df_mat_only <- tidyr::expand_grid(
  .draw = seq_len(nrow(fec_draws_df)),
  mat_id = 1:4
) %>%
  mutate(
    diff_rel_minus_unrel =
      exp(alpha_fix_rel + mat_df_rel[cbind(.draw, mat_id)]) -
      exp(alpha_fix_unrel + mat_df_unrel[cbind(.draw, mat_id)])
  ) %>%
  left_join(mat_key_fec_plot, by = "mat_id") %>%
  mutate(
    mat_label = factor(mat_label, levels = plot_mat_levels),
    plot_label = as.character(mat_label)
  )

# Overall average across maternal environments, computed within posterior draw
df_mat_overall <- df_mat_only %>%
  group_by(.draw) %>%
  summarise(
    diff_rel_minus_unrel = mean(diff_rel_minus_unrel),
    plot_label = "Overall",
    .groups = "drop"
  )

# Numeric x positions to create a gap before Overall
plot_x_positions <- c(
  "AN" = 1,
  "NN" = 2,
  "AA" = 3,
  "NA" = 4,
  "Overall" = 5.75
)

df_env_plot <- df_mat_only %>%
  mutate(
    x_pos = unname(plot_x_positions[plot_label])
  )

df_overall_plot <- df_mat_overall %>%
  mutate(
    x_pos = unname(plot_x_positions[plot_label])
  )

# Summaries for plotting
df_env_summ <- df_env_plot %>%
  group_by(plot_label, x_pos) %>%
  summarise(
    med = median(diff_rel_minus_unrel),
    lo = quantile(diff_rel_minus_unrel, fec_lo_p),
    hi = quantile(diff_rel_minus_unrel, fec_hi_p),
    .groups = "drop"
  )

df_overall_summ <- df_overall_plot %>%
  group_by(plot_label, x_pos) %>%
  summarise(
    med = median(diff_rel_minus_unrel),
    lo = quantile(diff_rel_minus_unrel, fec_lo_p),
    hi = quantile(diff_rel_minus_unrel, fec_hi_p),
    .groups = "drop"
  )

ggplot() +
  geom_hline(yintercept = 0, linewidth = 0.7) +

  # oxygen-environment-specific distributions
  ggdist::stat_halfeye(
    data = df_env_plot,
    aes(x = x_pos, y = diff_rel_minus_unrel),
    side = "right",
    justification = -0.15,
    width = 0.72,
    point_interval = NULL,
    alpha = 0.8,
    fill = contrast_color,
    slab_color = NA
  ) +

  # overall distribution, drawn separately so it can have its own width
  ggdist::stat_halfeye(
    data = df_overall_plot,
    aes(x = x_pos, y = diff_rel_minus_unrel),
    side = "right",
    justification = -0.15,
    width = 0.42,
    point_interval = NULL,
    alpha = 0.8,
    fill = contrast_color,
    slab_color = NA
  ) +

  geom_errorbar(
    data = df_env_summ,
    aes(x = x_pos, y = med, ymin = lo, ymax = hi),
    width = 0,
    linewidth = 0.7,
    color = "black"
  ) +
  geom_point(
    data = df_env_summ,
    aes(x = x_pos, y = med),
    size = 2.3,
    color = "black"
  ) +

  geom_errorbar(
    data = df_overall_summ,
    aes(x = x_pos, y = med, ymin = lo, ymax = hi),
    width = 0,
    linewidth = 0.7,
    color = "black"
  ) +
  geom_point(
    data = df_overall_summ,
    aes(x = x_pos, y = med),
    size = 2.3,
    color = "black"
  ) +

  geom_vline(
    xintercept = 4.875,
    linetype = "dotted",
    linewidth = 0.5
  ) +
  scale_x_continuous(
    breaks = unname(plot_x_positions),
    labels = names(plot_x_positions),
    limits = c(0.5, 6.25)
  ) +
  coord_cartesian(ylim = c(-3, 3)) +
  labs(
    x = axis_label_mat_env_full,
    y = "Light-Plasticity Fecundity Differentiation"
  ) +
  growth_base_plot_theme()
```



```{r fec_plot_difference_anc_vs_rel_unrel_intercept_only, echo=FALSE, message=FALSE, warning=FALSE}
alpha_anc <- fec_draws_df$`alpha[1]` # ancestral intercept
alpha_rel <- fec_draws_df$`alpha[2]` # reliable intercept
alpha_unrel <- fec_draws_df$`alpha[3]` # unreliable intercept

# calc change in fecundity due to intercept differences only
df_draws_rel <- tibble(
  .draw = seq_len(nrow(fec_draws_df)),
  diff_fit = exp(alpha_rel) - exp(alpha_anc),
  contrast = factor(
    "Reliable - Ancestral",
    levels = c("Reliable - Ancestral", "Unreliable - Ancestral")
  ),
  x_dummy = factor("", levels = "")
)

df_draws_unrel <- tibble(
  .draw = seq_len(nrow(fec_draws_df)),
  diff_fit = exp(alpha_unrel) - exp(alpha_anc),
  contrast = factor(
    "Unreliable - Ancestral",
    levels = c("Reliable - Ancestral", "Unreliable - Ancestral")
  ),
  x_dummy = factor("", levels = "")
)

# combine Rel and Unrel dfs
df_draws_both <- bind_rows(df_draws_rel, df_draws_unrel)

# summarise median and quantiles
df_summ_both <- df_draws_both %>%
  group_by(contrast, x_dummy) %>%
  summarise(
    med = median(diff_fit),
    lo = quantile(diff_fit, fec_lo_p),
    hi = quantile(diff_fit, fec_hi_p),
    .groups = "drop"
  )

# display plot
growth_plot_halfeye_with_summary(
  df_draws = df_draws_both,
  df_summ = df_summ_both,
  x = "x_dummy",
  y = "diff_fit",
  x_label = NULL,
  y_label = "Fecundity Divergence",
  remove_x_text = TRUE
) +
  labs(x = NULL) +
  coord_cartesian(ylim = c(-4,4)) + 
  facet_wrap(~ contrast, nrow = 1) +
  boxed_strip_theme(12) +
  theme(
    legend.position = "none"
  )+ labs(x = "Experimental Evolution Contrast")
```


```{r fec_plot_fecundity_variance_pairs_shared_pop, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=8}
xlim_common <- c(0, 1.2)
ylim_common <- c(0, 1.2)

draws_var <- fec_draws_df %>%
  transmute(
    `Block SD` = sigma_block,
    `Replicate Population SD` = sigma_pop,
    `Likelihood Residual SD` = 1 /phi
  )

draw_variance_matrix <- function(df_plot) {
  vars <- names(df_plot)
  k <- length(vars)

  grid::grid.newpage()
  grid::pushViewport(grid::viewport(layout = grid::grid.layout(k, k)))

  for (i in seq_len(k)) {
    for (j in seq_len(k)) {
      grid::pushViewport(
        grid::viewport(layout.pos.row = i, layout.pos.col = j)
      )

      xvar <- vars[j]
      yvar <- vars[i]

      if (i == j) {
        p <- ggplot(df_plot, aes(x = .data[[xvar]])) +
          geom_density(fill = "steelblue", alpha = 0.7, color = "navy") +
          coord_cartesian(xlim = xlim_common) +
          labs(x = xvar, y = "Density") +scale_x_continuous(n.breaks = 3)+
scale_y_continuous(n.breaks = 3)+
          theme_bw(base_size = 12) +
          theme(
            axis.title.x = element_text(size = 12),
  axis.title.y = element_text(size = 12),
  axis.text.x  = element_text(size = 8),
  axis.text.y  = element_text(size = 8)
          )  
      } else {
        p <- ggplot(df_plot, aes(x = .data[[xvar]], y = .data[[yvar]])) +
          geom_point(size = 0.4, alpha = 0.5, color = "navy") +
          coord_cartesian(xlim = xlim_common, ylim = ylim_common) +
          labs(x = xvar, y = yvar) +scale_x_continuous(n.breaks = 3)+
scale_y_continuous(n.breaks = 3)+
          theme_bw(base_size = 12) +
          theme(
            axis.title.x = element_text(size = 12),
  axis.title.y = element_text(size = 12),
  axis.text.x  = element_text(size = 8),
  axis.text.y  = element_text(size = 8)
          )    }

      print(p, vp = grid::viewport())
      grid::popViewport()
    }
  }

  grid::popViewport()
}

draw_variance_matrix(draws_var)
```
